% Representational Incompleteness — LaTeX source (Lean 4 / Mathlib companion library)

\documentclass[11pt]{article}
\input{nova_zenodo_doi_placeholder}
\usepackage[margin=1in]{geometry}
\usepackage{amsmath,amssymb,amsthm}
\usepackage{hyperref}
\usepackage{enumitem}
\usepackage{array}
\usepackage{booktabs}
\usepackage{longtable}
\usepackage{xcolor}
\usepackage{stmaryrd}
\usepackage{tcolorbox}
\newtcolorbox{keybox}[1][]{
  colback=blue!4, colframe=blue!35,
  arc=3pt, boxrule=0.5pt,
  fonttitle=\bfseries, #1
}

% Allow line breaks in long Lean identifiers (\_, namespaces, CamelCase).
\ExplSyntaxOn
\tl_new:N \l__rr_leanref_tl
\NewDocumentCommand \leanref { m }
  {
    \tl_set:Nn \l__rr_leanref_tl { #1 }
    \regex_replace_all:nnN { \c{_} } { \c{_} \c{allowbreak} } \l__rr_leanref_tl
    \regex_replace_all:nnN { \. } { . \c{allowbreak} } \l__rr_leanref_tl
    % CamelCase / digit-boundary splits (Lean module and declaration names).
    \regex_replace_all:nnN { ([a-z])([A-Z]) } { \1 \c{allowbreak} \2 } \l__rr_leanref_tl
    \regex_replace_all:nnN { ([0-9])([A-Z]) } { \1 \c{allowbreak} \2 } \l__rr_leanref_tl
    \texttt{\small \tl_use:N \l__rr_leanref_tl }
  }
\ExplSyntaxOff

\newtheorem{theorem}{Theorem}[section]
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{proposition}[theorem]{Proposition}
\newtheorem{corollary}[theorem]{Corollary}
\newtheorem{definition}[theorem]{Definition}
\newtheorem{remark}[theorem]{Remark}

\newcommand{\novel}{\textcolor{blue!70!black}{[Novel]}}
\newcommand{\known}{\textcolor{gray}{[Known math]}}
% Break `theorem`/`definition` head away from long \leanref tags (avoids overfull italic headers).
\newcommand{\TheoremLeanTag}[1]{%
  \noindent\textnormal{(\leanref{#1})}\par\noindent}
% End `amsthm` head paragraph so the declaration tag is not merged into the header line.
\newcommand{\AfterTheoremHead}{\mbox{}\par}

\title{Representational Incompleteness:\\
  No Self-Modeling System Can Capture Its Own Diagonal\\[0.3em]
  {\large A Machine-Checked Barrier with Six Closed Escape Routes}}
\author{Nova Spivack\\
Independent Researcher}
\date{2026}

\begin{document}
\maketitle

% ============================================================================
% ABSTRACT
% ============================================================================
\begin{abstract}
\raggedright
\noindent\textbf{Summit theorem (\leanref{representational\_incompleteness}).}
Any system that models itself---any $s : A \to A \to B$ whatsoever---is
\emph{necessarily incomplete} with respect to its own diagonal behavior:
the function $a \mapsto f(s\,a\,a)$ is never equal to any row $s\,a_0$
of the self-model, for every fixed-point-free $f : B \to B$ and every
$a_0 \in A$.  No surjectivity, totality, or cardinality assumption on
$s$ is required.  If $s$ is claimed to be total/surjective, it simply
cannot exist.

This is \textbf{representational incompleteness}: the analogue of
G\"odel incompleteness, not for formal theories and provability, but
for any system that models itself via a parametric family of functions.
The blind spot is structural---not a resource limitation---and
persists for every codomain admitting a fixed-point-free endomorphism
(including $\mathbb{N}$ with successor and $\mathrm{Bool}$ with
negation).
Unlike G\"odel incompleteness, representational incompleteness
requires no formal theory, no arithmetic, and no coding; it holds for
every parametric self-model regardless of computability, cardinality,
or totality.

\medskip
\noindent\textbf{Topological reinforcement.}
A second, independent strand proves that the barrier is not merely
diagonal but \emph{topological}: a general lemma states that any two
spaces whose intrinsic planar-chart boundaries are subtype-incompatible
are not homeomorphic.  The obstacle is \textbf{wrong type}, not
\textbf{inexhaustible queue}---a finite-data invariant that no amount
of computation, scaling, or architectural extension can change.  The
lemma applies to an \emph{open-ended class} of target spaces (M\"obius
strip, Klein bottle, projective plane, or any non-orientable surface
with incompatible boundary); the closed cylinder and M\"obius strip are
the minimal fully discharged compact witness.

\medskip
\noindent\textbf{No escape.}
Six adversarial retreat routes are formally closed: infinite regress,
finite-order iterate collapse, cross-object readout (both sides),
partial self-models, decidable self-reference, and model substitution.
Each is a proved theorem in Lean~4, checked against Mathlib, with zero
\texttt{sorry} and no axioms beyond Lean~4's standard foundation
(propositional extensionality, quotients, choice).

\medskip
\noindent\textbf{Three pillars.}
The result rests on three mutually reinforcing pillars:
(1)~\emph{diagonal exclusion}---every self-model has a blind spot
shaped like its own diagonal;
(2)~\emph{topological obstruction}---a general lemma defining an
open-ended class of space-pairs that are not homeomorphic, detected by
a finite-data boundary invariant impervious to computation;
(3)~\emph{ontological sort wall}---the \texttt{OntologicalSlot}
construction enforces that the self-representation morphism and any
global element remain definitionally disjoint at every iterate depth;
the disjointness is proved robust under arbitrary injective relabeling
(\leanref{ClosureProperties}), so it is a structural invariant, not a
notational artifact.

\medskip
\noindent\textbf{Closure properties.}
The diagonal exclusion is proved robust under composition,
reindexing, post-composition, embedding, product formation,
multi-agent delegation, and nested simulation.  A formal
\emph{bridge hypothesis}~P3 makes the modeling claim ``the system
has a parametric self-model'' explicit and conditional, parallel
to~P2 for consciousness.  A \leanref{TopologicalSelfModelingSystem}
structure formally bridges Pillars~1 and~2.

\medskip
\noindent\textbf{Consequence.}
In every neighboring debate---consciousness, symbolic AI, semantic
grounding, measurement chains, LLM scaling, AGI self-improvement, and
the simulation hypothesis---the six closed escape routes
leave every remaining admissible path requiring \emph{structural change}
to the topological type of the substrate, not quantitative extension
within a fixed type.  Scaling, iterating, partial modeling, cross-object
readout, finite-depth loops, and decidable restriction are all formally
inadmissible.

\medskip
\noindent\textbf{What fills the blind spot.}
The blind spot is located at the self-referential diagonal.  A
functioning self-modeling system \emph{does} produce diagonal
behavior, but its self-model \emph{cannot} represent that behavior.
Therefore the substrate---the underlying machinery executing the
system---is the \textbf{formally necessary complement} to any
parametric self-model.  No self-improvement, delegation, or scaling
can substitute for it.  Self-awareness with perfect self-transparency
is the one thing that is structurally impossible; self-awareness with
a diagonal opacity is what the formalism predicts and what empirical
evidence confirms.

\medskip
\noindent\textbf{Trust boundary:} Diagonal and topological clauses are Lean theorems at the cited declarations; philosophical deployment claims require explicit modeling bridges beyond the formal artifact.

\NovaZenodoInAbstract
\end{abstract}

\noindent\textbf{Keywords:} representational incompleteness, self-reference, Lean~4, diagonal arguments, topology

\noindent\textbf{MSC 2020:} 03F40, 03B70, 54A05, 68T27

\noindent\textbf{Code/Data availability:}
\texttt{representational-incompleteness} Lean library;
\texttt{lake build};
repository \texttt{README} and module index at repository root.

\section{Premises, disagreement coordinates, and trust boundary}\label{sec:premises-ri-flagship}
\begin{keybox}[title=Where disagreement can (and cannot) live]
\textbf{Formal spine.}
The summit barrier and the six closed escape routes are discharged as \texttt{theorem} values in the artifact; informal debates route through those declarations.

\medskip
\textbf{Premise coordinates (for external readers).}
\begin{itemize}[leftmargin=2em]
\item \textbf{Generality of $s$.} The main diagonal exclusion assumes only a parametric self-map $s:A\to A\to B$ and a fixed-point-free $f$ on $B$; no computability or cardinality strengthening is required for the core lemma.
  \emph{Logical effect:} Adding resource bounds is an optional specialization, not the negation of the base theorem.
\item \textbf{Topological strand.} The planar-chart boundary lemma is a separate pillar with its own hypotheses; it reinforces but does not silently replace the diagonal clause.
  \emph{Logical effect:} One may accept diagonal incompleteness while auditing topology proofs independently.
\end{itemize}
\end{keybox}

% ============================================================================
% 1. INTRODUCTION
% ============================================================================
\section{Introduction}
\label{sec:intro}

\subsection{The Problem and the Summit Theorem}
When a system models itself---by iterating a fixed construction, by
encoding its own states, by interpreting its own symbols---a familiar worry
arises: infinite regress, or diagonal limitations, or both.  But those
concerns are usually framed \emph{computationally}: the regress does not
terminate, the self-interpreter cannot be total, the G\"odel sentence is
undecidable.  The implicit hope is that with enough resources---more
compute, richer architectures, deeper self-models---the obstacle might
shrink or be routed around.

This paper proves it cannot.  The headline result is:

\begin{theorem}[\novel\ Representational Incompleteness]
\label{thm:summit}
\AfterTheoremHead
\TheoremLeanTag{representational\_incompleteness}
Let $B$ be a type with a fixed-point-free endomorphism $f : B \to B$.
For \textbf{any} $s : A \to A \to B$ (no surjectivity, totality, or
cardinality assumption):
\begin{enumerate}[label=(\alph*)]
\item \textbf{Diagonal exclusion:}
  for every $a_0 \in A$,
  $\;s\,a_0 \neq \lambda a.\,f(s\,a\,a)$.
\item \textbf{Total coverage impossible:}
  $\neg\!\left(\forall g : A \to B,\;\exists a,\;s\,a = g\right)$.
\end{enumerate}
\end{theorem}

\noindent
Every self-modeling system is \emph{necessarily incomplete}: its own
diagonal behavior is precisely the information it cannot represent.  If
$s$ is claimed to be total and surjective, it simply cannot exist.  This
is \textbf{representational incompleteness}---the analogue of G\"odel
incompleteness, not for formal theories and provability, but for any
system that models itself via a parametric family of functions.  The
blind spot is structural, not a resource limitation, and persists for
$B = \mathbb{N}$ (successor), $B = \mathrm{Bool}$ (negation), and every
other codomain admitting a fixed-point-free endomorphism.

A functioning self-modeling system \emph{does} produce diagonal
behavior---it processes self-referential inputs and generates
outputs.  Since no row of its own self-model matches that behavior,
the diagonal behavior must be generated by the underlying
\emph{substrate}: the physical or computational machinery executing
the system.  The substrate is the formally necessary complement to
any parametric self-model (\S\ref{sec:substrate-inference}).

A second, independent strand reinforces this with a
\textbf{topological} barrier: a general lemma showing that
\emph{any} two spaces whose intrinsic planar-chart boundaries
are subtype-incompatible are not homeomorphic---an open-ended
class of obstructions that no amount of iteration, scaling, or
continuous deformation within a fixed topological type can
remove.  And we close, one by one, every standard adversarial
route by which one might hope to circumvent either barrier.

The technical development is in Lean~4, checked against Mathlib, with
zero \texttt{sorry} and no axioms beyond Lean~4's standard foundation;
\S\ref{sec:instantiations}--\S\ref{sec:admissibility} relate the
results to consciousness, symbolic grounding, diagonal arguments,
the simulation hypothesis, AI/AGI scaling, and observer--apparatus
topologies.

\begin{keybox}[title=This is not G\"odel repackaged]
Representational incompleteness differs from G\"odel incompleteness in
every structural respect.  G\"odel requires a formal theory rich enough
for arithmetic, a G\"odel coding, and a self-referential
\emph{sentence}; the blind spot is an \emph{unprovable truth}.
Representational incompleteness requires only a type $B$ with a
fixed-point-free endomorphism and \emph{any} $s : A \to A \to B$; the
blind spot is an \emph{unrepresentable function}---the system's own
diagonal.  No formal theory, no arithmetic, no coding, no
computability assumption is needed.  The result is \textbf{stronger}
in one respect: it holds for every $s$ regardless of whether $s$ is
computable, whereas G\"odel- and Turing-style arguments require the
system to be recursively axiomatizable or computable.
\end{keybox}

\subsection{Summary of Results}

The core results are \textbf{representational incompleteness}
(Theorem~\ref{thm:summit}) and a \textbf{universal topological
obstruction}---reinforced by a proof that the six most natural
adversarial attempts to evade them all fail.

\paragraph{The topological obstruction.}
A general lemma (Theorem~\ref{thm:generic}) states: if two spaces admit
subsets characterized as exactly the non-\texttt{ChartableR2} points, and
if those boundary \emph{subspaces} are not homeomorphic, then the ambient
spaces are not homeomorphic.  This is a \emph{finite-data invariant}: two
spaces are either homeomorphic or not, regardless of how many modeling
steps one enumerates.  The lemma defines an \emph{open-ended class} of
obstructed pairs---any two spaces meeting the three hypotheses, regardless
of genus, orientability, or dimension---rather than a single hard-wired
example.  The closed cylinder and M\"obius strip serve as the
\textbf{minimal-case witness} (\textbf{M-FINAL}); Klein bottles,
projective planes, higher-genus non-orientable surfaces, etc., would
equally suffice.  Any future domain whose state space matches the same
boundary pattern imports the same conclusion with no new proof.

\paragraph{Why this is not ``just'' infinite regress---or ``just'' G\"odel.}
The categorical strand yields infinitely many distinct meta-levels, but
infinity is not the deepest point.  The topological obstruction is
independent: it says the barrier is \textbf{wrong type}, not
\textbf{inexhaustible queue}.  Nor is this ``just G\"odel'':
representational incompleteness requires no formal theory, no arithmetic,
no coding, and no computability assumption on~$s$.  The blind spot is a
\emph{function}, not a sentence, and no theory extension can close it.
Section~\ref{sec:godel-distinction} provides a detailed comparison
on six independent structural axes.

\begin{keybox}[title=The topological barrier is independent of diagonalization]
The topological strand (Theorem~\ref{thm:generic}) is
\emph{completely orthogonal} to Cantor, G\"odel, Lawvere, and every
other diagonal argument.  It detects a \emph{finite-data invariant}:
the connectivity pattern of the non-chartable boundary.  No amount
of iteration, scaling, architectural reparameterization, or
continuous deformation within a fixed topological type can alter
this invariant.  The obstacle is \textbf{wrong type}, not
\textbf{inexhaustible queue}---and the general lemma defines an
\emph{open-ended class} of obstructed pairs, not a single hard-wired
example.
\end{keybox}

\paragraph{No escape.}
Six adversarial retreat routes---infinite regress, finite-order
collapse, cross-object readout, partial self-models, decidable
self-reference, and model substitution---are each formally closed by
a proved theorem (\S\ref{sec:noescape}).  Together they leave an
adversary with a clean dichotomy: either an infinite injective
regress with the Lawvere wall intact, or a finite loop too shallow
for full self-capture---and in either case the topological barrier
stands.

\paragraph{Bridges and conditional readings.}
Cross-domain structures
(\leanref{SymbolSystem},
\leanref{LawvereRegressBridge},
bundled \leanref{ChartableR2BoundaryModel} instances,
\leanref{QuantumObserverChainHypothesis})
make the obstruction's \emph{hypothesis shape} explicit for grounding,
diagonal, consciousness~(under premise~P2), and observer--readout
readings.  Because the topological obstruction is a general lemma over
a class of space-pairs, each bridge inherits the full generality
automatically.  An extended master theorem adjoins all three layers
(diagonal, topological, categorical) in one checkable bundle.

\paragraph{Utility.}
For any application the workflow is:
(1)~encode the system's state space in sufficient topological detail to
verify the boundary biconditionals,
(2)~compare boundary subspaces up to homeomorphism,
(3)~invoke the general lemma or a bundled model.
If the premises hold, the obstruction applies; if they fail, the
obstacle was never licensed.  That conditional structure is what keeps
the mathematics honest while delivering a \emph{portable} gate.

\subsection{Contributions and Novelty Claims}

Individual ingredients are classical: Lawvere's diagonal (1969), textbook
topology of surfaces, Cantorian regress patterns.
The contribution is a \textbf{unified, adversarially checked, machine-verified}
landscape in which representational incompleteness, topological obstruction,
and six closed escape routes coexist in one artifact, so that ``compute
harder'' is never confusable with ``change homeomorphism type.''  Specific
novel elements:

\begin{enumerate}[label=(\arabic*),itemsep=0.4ex]
\item \textbf{\texttt{OntologicalSlot}} --- inductive tagging of morphism
  vs.\ global-element slots; makes the Lawvere wall explicit inside the
  type discipline. [\novel]
\item \textbf{\texttt{ChartableR2} spine + seam--equator charts} ---
  intrinsic $\mathbb{R}^2$ chartability as a definable boundary sensor;
  constructive charts through the M\"obius quotient seam including the
  equator via piecewise unfolding.  The M\"obius strip serves as the
  minimal proof witness; the predicate applies uniformly to all
  surfaces. [\novel\ technique]
\item \textbf{General boundary-contrast obstruction} ---
  a single reusable, machine-checked lemma defining an
  \emph{open-ended class} of obstructed pairs: any two spaces
  with incompatible \leanref{ChartableR2}-co-characterized
  boundaries are not homeomorphic---regardless of genus,
  orientability, or dimension.  To our knowledge, \emph{not}
  standard in the philosophy-of-mind or AI-safety
  literatures. [\novel]
\item \textbf{Cross-domain bridge structures} ---
  \texttt{SymbolSystem}, \texttt{LawvereRegressBridge},
  bundled \texttt{ChartableR2BoundaryModel} data,
  \texttt{QuantumObserverChainHypothesis}: these make the obstruction's
  hypothesis shape explicit per domain. [\novel]
\item \textbf{No-escape horn closures} --- six formally blocked
  adversarial escape routes (iterate collision, section--retraction on
  both sides, partial-model diagonal exclusion, decidable Lawvere,
  wall persistence at every depth).  The integrated closure package is,
  to our knowledge, new. [\novel]
\item \textbf{Closure properties and bridge hypothesis P3} ---
  compositional inheritance of the diagonal exclusion through
  post-composition, reindexing, embedding, and product formation;
  type-level cross-object readout; a formal \texttt{ComputationalSelfModel}
  bridge (P3) making the AI/computational modeling claim explicit and
  conditional; slot robustness theorems proving the endo/obj wall
  survives any injective relabeling; per-pillar full-closure witnesses;
  and a \texttt{TopologicalSelfModelingSystem} structure formally
  bridging Pillars~1 and~2. [\novel]
\end{enumerate}

\subsection{Paper Organization}
The paper is structured around three layers that build toward a single
conclusion.

\textbf{Layer~1: Diagonal \& categorical}
(\S\ref{sec:regress})---representational systems, regress, Lawvere wall,
\texttt{OntologicalSlot}, iterate dichotomy, core master theorem.

\textbf{Layer~2: Topological obstruction}
(\S\S\ref{sec:models}--\ref{sec:mfinal})---minimal-case witnesses
(cylinder, M\"obius strip), \texttt{ChartableR2} boundary sensor, general
obstruction lemma, seam charts, M-FINAL.

\textbf{Layer~3: Integration \& closure}
(\S\ref{sec:extended})---extended master theorem, instantiations
(consciousness, grounding, diagonal arguments, QM, simulation
hypothesis, LLM scaling, AI/AGI), anatomy of the blind
spot~(\S\ref{sec:blind-spot-anatomy}),
horn-by-horn no-escape analysis~(\S\ref{sec:noescape}),
admissibility~(\S\ref{sec:admissibility}).

\S\S\ref{sec:limits}--\ref{sec:conclusion}: limitations, related work,
conclusion.  Appendices: Lean module map, key theorem index, reproduction.

% ============================================================================
% 2. CATEGORICAL STRAND: THE REPRESENTATIONAL REGRESS
% ============================================================================
\section{The Representational Regress}
\label{sec:regress}

This section builds the categorical machinery that feeds into the summit
theorem: the endomorphism tower, the Lawvere wall, and the
\texttt{OntologicalSlot} distinction.  Together they establish that
self-representation via iterated endomorphisms either produces an
infinite regress with the endo/obj wall intact, or collapses to a finite
loop too shallow for injective self-capture---setting the stage for the
diagonal exclusion of Theorem~\ref{thm:summit}.

\subsection{The \texttt{RepresentationalSystem} Structure}
\label{sec:repsys}

\begin{definition}[\leanref{Basic.lean}: \leanref{RepresentationalSystem}]
\leavevmode\par\noindent
A \emph{representational system} is a tuple $(C, A, \mathrm{represent}, \mathrm{iter\_injective})$ where:
\begin{itemize}
\item $C$ is a category,
\item $A$ is an object of $C$,
\item $\mathrm{represent} : A \to A$ is an endomorphism,
\item $\mathrm{iter\_injective}$: for all $n \neq m$, $\mathrm{represent}^n \neq \mathrm{represent}^m$ in $\mathrm{End}(A)$.
\end{itemize}
\end{definition}

\subsection{Non-Termination and Unboundedness}

\begin{theorem}[\known\ \leanref{Regress.lean}]
\label{thm:regress}
For any representational system $R$:
\begin{enumerate}[label=(\alph*)]
\item (\leanref{regress\_no\_termination}) Distinct indices $n \neq m$ yield distinct meta-level arrows.
\item (\leanref{regress\_iterates\_unbounded}) For every bound $B$, there exists a level $L > B$ with a distinct iterate.
\item (\leanref{regress\_is\_infinite}, \leanref{meta\_range\_infinite}) The set of meta-levels is infinite.
\end{enumerate}
\end{theorem}

\begin{remark}[What if \texttt{iter\_injective} fails?]
\label{rem:dichotomy}
The hypothesis $\mathrm{iter\_injective}$ is \emph{bundled}, not derivable from
a bare endomorphism. An adversary might weaken it: suppose
$\mathrm{represent}^n = \mathrm{represent}^m$ for some $n \neq m$.
\leanref{NoEscapeClosure} formalizes the consequence: let
$e := \mathrm{End.of}(\mathrm{represent})$; if $e^{n+d}=e^n$ with $d>0$ then
every power $e^k$ equals some $e^i$ with $i < n+d$
(\leanref{IterateClosure.End.exists\_lt\_of\_pow\_collision}).
\leanref{IterateClosure.End.pow\_iterate\_dichotomy}
records the \textbf{dichotomy}: for every monoid element~$e$, either
(i)~distinct natural-number powers are all distinct (an infinite injective
$\mathbb{N}$-tower), or (ii)~the set of \emph{distinct} powers is bounded by a
finite index.  In case~(ii) the system has only finitely many representational
layers---it loops.  Either route blocks ``just enough'' self-representation: one
gives an infinite regress; the other gives a finite loop too impoverished to
host injective iterate depth.
\end{remark}

\subsection{The Lawvere Wall}
\label{sec:lawvere}

\begin{theorem}[\known\ \leanref{Lawvere.lean}]
\label{thm:lawvere}
(\leanref{lawvere\_fixed\_point\_Type}) If $s : A \to (A \to B)$ is surjective, then every $f : B \to B$ has a fixed point.
\end{theorem}

\begin{theorem}[\known\ \leanref{LawvereCCCType.lean}]
\label{thm:lawvere-ccc}
(\leanref{lawvere\_fixed\_point\_MonoidalClosedType}) The same conclusion holds when the universal-enumerator hypothesis is recast as surjectivity of $\mathrm{curry}(\mathrm{lawvereBinary}\ s)$ in $\mathrm{MonoidalClosed}(\mathrm{Type}\ u)$.
\end{theorem}

\subsection{The \texttt{OntologicalSlot} Construction}
\label{sec:ontoslot}

\begin{definition}[\novel\ \leanref{Basic.lean}: \leanref{OntologicalSlot}]
\label{def:ontoslot}
For a category $C$ with terminal object and an object $A$:
\[
\mathrm{OntologicalSlot}(C, A) \;:=\;
  \mathrm{endo}(f : A \to A) \;\mid\; \mathrm{obj}(g : \top_C \to A)
\]
an inductive type with two constructors, disjoint by definition.
\end{definition}

\begin{theorem}[\novel\ \leanref{FixedPoints.lean}, \leanref{LawvereRegressBridge.lean}]
\label{thm:wall}
\leavevmode
\begin{enumerate}[label=(\alph*)]
\item (\leanref{lawvere\_wall\_is\_not\_dissolution}) For any representational system $R$ and any global element $\mathrm{fp} : \top_C \to A$,
$\;\mathrm{OntologicalSlot.endo}(R.\mathrm{represent}) \neq \mathrm{OntologicalSlot.obj}(\mathrm{fp}).$
\item (\leanref{lawvere\_fixed\_point\_stays\_representational}) The wall holds at \textbf{every iterate depth}: for all $n : \mathbb{N}$ and all $c : C$,
$\;\mathrm{OntologicalSlot.endo}(\mathrm{represent}^n) \neq \mathrm{OntologicalSlot.obj}(c).$
\end{enumerate}
\end{theorem}

\subsection{Why \texttt{OntologicalSlot} Is Novel: The Wall That Lawvere Does Not Build}
\label{sec:ontoslot-novelty}

Lawvere's fixed-point theorem (1969) proves that if $s : A \to (A \to B)$
is surjective, then every $f : B \to B$ has a fixed point.  This is a
theorem about surjectivity and fixed points.  It does \emph{not} tag the
fixed point as belonging to a different ontological sort from the
morphism that produced it.

In standard category theory, the endomorphism monoid $\mathrm{End}(A)$
and the set of global elements $\mathrm{Hom}(\mathbf{1}, A)$ are both
sets of morphisms in the ambient category.  There is no formal marker
distinguishing ``this morphism is a process acting on $A$'' from ``this
morphism is a state of $A$ selected from the terminal object.''  The two
live in different hom-sets, but no inductive tag enforces their
disjointness inside a single data structure.

\texttt{OntologicalSlot} (Definition~\ref{def:ontoslot}) fills this gap.
It is an inductive type with two constructors---$\mathrm{endo}(f)$ and
$\mathrm{obj}(g)$---that are \emph{definitionally disjoint} by
constructor discrimination.  The wall theorem
(Theorem~\ref{thm:wall}) then proves that for any representational
system, the self-representation morphism $\mathrm{represent}$ (tagged as
$\mathrm{endo}$) is never equal to any global element $\mathrm{fp}$
(tagged as $\mathrm{obj}$), regardless of how many times
$\mathrm{represent}$ is iterated.  The Lawvere fixed point exists as an
object-witness; it does \emph{not} dissolve the endo/obj wall.

\paragraph{Philosophical significance.}
In philosophy of mind, the intuition that ``the \emph{process} of
self-representation and the \emph{product} of self-representation inhabit
different ontological categories'' is widespread but usually left
informal.  \texttt{OntologicalSlot} is its machine-checked
formalization.  The earlier ``Transputation'' argument (Spivack, 2025)
claimed that in the base case of awareness aware of itself ($A \to A$),
the roles of knower, known, and knowing ``collapse''---the process of
being aware and the state of being aware cannot be distinguished
phenomenologically.  \texttt{OntologicalSlot} formalizes the
\emph{impossibility} of that collapse within any categorical
representation: the iterate $\mathrm{represent}^n$ (a morphism, an
$\mathrm{endo}$) can never equal a global element $\mathrm{fp} :
\mathbf{1} \to A$ (a state, an $\mathrm{obj}$).  If consciousness
involves a genuine identification of process and product, then no
categorical self-model can achieve it.

\begin{keybox}[title=The wall that Lawvere does not build]
Lawvere's theorem guarantees a fixed point.
\texttt{OntologicalSlot} guarantees that the fixed point remains an
\emph{object}-witness, never collapsing into the
\emph{morphism}-process that generated it.  The ``wall'' between
$\mathrm{endo}$ and $\mathrm{obj}$ is a definitional invariant that
persists at every iterate depth---not proved by Lawvere, not present
in standard category theory, and not available in any prior
formalization of self-reference.
\end{keybox}

\paragraph{Robustness: the wall is structural, not notational.}
An adversary might object: ``The endo/obj disjointness is trivially
true by constructor discrimination---it is a notational convention,
not a mathematical barrier.''  This objection is addressed by the
\emph{robustness theorems} in \leanref{ClosureProperties.lean}:
\begin{itemize}[leftmargin=1.5em,itemsep=0.15ex,topsep=0.4ex,parsep=0pt]
\item \leanref{OntologicalSlot.wall\_survives\_injection}: for any
  injective relabeling $\varphi : \mathrm{OntologicalSlot} \to T$,
  the wall persists: $\varphi(\mathrm{endo}(f)) \neq \varphi(\mathrm{obj}(c))$.
\item \leanref{OntologicalSlot.wall\_collapse\_requires\_noninjective}:
  the \emph{only} way to collapse endo and obj into the same image is
  via a non-injective map---one that deliberately identifies distinct things.
\item \leanref{OntologicalSlot.iterate\_tower\_injective}: distinct
  iteration depths $n \neq m$ yield distinct endo-slots, so the iterate
  tower is faithfully represented in slot space.
\end{itemize}
The wall is therefore a \emph{structural invariant of the type
discipline}:  it survives any relabeling that respects distinctness.
To erase it, one must use a function that conflates distinct objects
by design---which is not an escape but a concession.

\subsection{The Core Master Theorem}

\begin{theorem}[\leanref{Main.lean}]
\label{thm:master}
(\leanref{representational\_regress\_master}) For any representational system $R$:
\[
  \mathrm{representational\_regress\_master\_claim}(R)
\]
i.e., the conjunction of unbounded regress, Lawvere wall non-dissolution, and $T_2$ invariance under homeomorphism.
\end{theorem}

% ============================================================================
% 3. TOPOLOGICAL MODELS
% ============================================================================
\section{Topological Models: Minimal-Case Witnesses}
\label{sec:models}

The general obstruction (Theorem~\ref{thm:generic}) applies to
any pair of spaces with incompatible
\leanref{ChartableR2}-detected boundaries.  This section
establishes the closed cylinder and M\"obius strip as the
\textbf{simplest fully discharged witness pair}: two compact
surfaces whose boundary subtypes are topologically incompatible.

\subsection{The Closed Cylinder}
\label{sec:cylinder}

\begin{definition}[\leanref{CylinderBoundary.lean}]
$\mathrm{ClosedCylinder} := S^1 \times [0,1]$ with the product topology and manifold structure $\mathfrak{R}_1 \times \mathfrak{R}_1^\partial$.
\end{definition}

\begin{theorem}[\known\ \leanref{CylinderBoundary.lean}]
\label{thm:cyl-boundary}
The manifold boundary of \emph{ClosedCylinder} decomposes as:
\[
  \partial(\mathrm{ClosedCylinder}) = F_{\mathrm{bot}} \sqcup F_{\mathrm{top}}
\]
where $F_{\mathrm{bot}} = S^1 \times \{0\}$ and $F_{\mathrm{top}} = S^1 \times \{1\}$ are each connected, and their union is \emph{disconnected}.
\end{theorem}

\subsection{The M\"obius Strip}
\label{sec:mobius}

\begin{definition}[\leanref{CylinderMobius.lean}]
\label{def:mobius-rel}
The \emph{M\"obius equivalence} $\sim$ on $[0,1]^2$ is the equivalence relation generated by
$(0, t) \sim (1, 1-t)$ for $t \in [0,1]$, extended by reflexivity, symmetry, and transitivity.
The \emph{M\"obius strip} is $\mathrm{MobiusStrip} := [0,1]^2 / {\sim}$.
\end{definition}

\begin{theorem}[\leanref{CylinderMobius.lean}]
\label{thm:mobius-props}
\leavevmode
\begin{enumerate}[label=(\alph*)]
\item (\leanref{mobiusRel0Graph\_isClosed}) The graph of $\sim$ is closed in $[0,1]^2 \times [0,1]^2$.
\item (\leanref{CompactSpace MobiusStrip}) $\mathrm{MobiusStrip}$ is compact (quotient of compact).
\item (\leanref{instT2SpaceMobiusStrip}) $\mathrm{MobiusStrip}$ is Hausdorff (closed graph + compact domain).
\item (\leanref{isConnected\_mobiusStripBoundary}) The boundary band $\partial_{\mathrm{model}} := \pi(\{t = 0\} \cup \{t = 1\})$ is connected (one component).
\end{enumerate}
\end{theorem}

\subsection{Boundary Subspace Contrast}

\begin{theorem}[\leanref{MobiusCylinderBoundaryContrast.lean}]
\label{thm:boundary-contrast}
\AfterTheoremHead
\TheoremLeanTag{mobiusStripBoundary\_not\_homeomorphic\_closedCylinderBoundaryUnion}
\[
  \mathrm{IsEmpty}\!\left(\mathrm{Subtype}(\partial_{\mathrm{Mob}}) \simeq_t \mathrm{Subtype}(\partial_{\mathrm{Cyl}})\right).
\]
The M\"obius boundary band (connected) and the cylinder boundary union (disconnected) are not homeomorphic as subspaces.
\end{theorem}

% ============================================================================
% 4. THE CHARTABLER2 FRAMEWORK
% ============================================================================
\section{The \texttt{ChartableR2} Framework}
\label{sec:chartable}

The previous sections established diagonal and categorical barriers.
This section introduces the topological layer: a boundary sensor based on
intrinsic $\mathbb{R}^2$-chartability, together with the general
obstruction lemma (Theorem~\ref{thm:generic}) that will deliver the
class-level topological barrier.

\subsection{Definition and Basic Properties}

\begin{definition}[\leanref{ChartableR2Neighbor.lean}]
\label{def:chartable}
A point $x$ of a topological space $X$ is \emph{$\mathrm{ChartableR2}$} if there exists an open set $U \ni x$ such that $U \cong_t \mathbb{R}^2$.
\end{definition}

\begin{theorem}[\leanref{ChartableR2Neighbor.lean}]
\label{thm:chartable-props}
\leavevmode
\begin{enumerate}[label=(\alph*)]
\item (\leanref{isOpen\_setOf\_chartableR2}) The set of \texttt{ChartableR2} points is open.
\item (\leanref{isClosed\_setOf\_not\_chartableR2}) Its complement is closed.
\item (\leanref{not\_chartableR2\_of\_tendsto\_seq\_atTop}) Sequential stability: limits of non-chartable points are non-chartable.
\end{enumerate}
\end{theorem}

\subsection{Invariance Under Homeomorphism}

\begin{theorem}[\leanref{ChartableR2Bridge.lean}]
\label{thm:chartable-invar}
\AfterTheoremHead
\TheoremLeanTag{Homeomorph.chartableR2\_iff}
If $h : X \simeq_t Y$ is a homeomorphism, then $\mathrm{ChartableR2}(x) \iff \mathrm{ChartableR2}(h(x))$.
\end{theorem}

\subsection{The Generic Boundary-Contrast Obstruction}
\label{sec:generic}

\begin{theorem}[\novel\ \leanref{ChartableR2Bridge.lean}]
\label{thm:generic}
\AfterTheoremHead
\TheoremLeanTag{not\_homeomorphic\_of\_chartableR2\_boundary\_contrast}
Let $X$ and $Y$ be topological spaces with predicates $b_X \subseteq X$ and $b_Y \subseteq Y$ such that:
\begin{enumerate}[label=(\roman*)]
\item $\forall x,\; x \in b_X \iff \neg\,\mathrm{ChartableR2}(x)$,
\item $\forall y,\; y \in b_Y \iff \neg\,\mathrm{ChartableR2}(y)$,
\item $\mathrm{IsEmpty}(\mathrm{Subtype}(b_X) \simeq_t \mathrm{Subtype}(b_Y))$.
\end{enumerate}
Then $\mathrm{IsEmpty}(X \simeq_t Y)$.
\end{theorem}

\begin{proof}[Proof sketch]
Any homeomorphism $e : X \simeq_t Y$ preserves $\mathrm{ChartableR2}$
(Theorem~\ref{thm:chartable-invar}), hence maps $b_X$ to $b_Y$ pointwise.
By $e.\mathrm{subtype}$, we get a homeomorphism
$\mathrm{Subtype}(b_X) \simeq_t \mathrm{Subtype}(b_Y)$, contradicting~(iii).
\end{proof}

Theorem~\ref{thm:generic} subsumes the non-homeomorphism strategy of \S\ref{sec:mfinal}:
any other pair of spaces satisfying the same three hypotheses is also not homeomorphic.

\subsection{The Topological Barrier in Detail}
\label{sec:topo-barrier-detail}

Theorem~\ref{thm:generic} deserves sustained attention because it
establishes a barrier of a fundamentally different character from
anything in the Cantor--G\"odel--Lawvere tradition.  This subsection
unpacks what the barrier is, why it is impervious to computation, and
what its generality implies.

\paragraph{What \texttt{ChartableR2} detects.}
A point $x$ of a topological space $X$ is $\mathrm{ChartableR2}$ if it
has an open neighborhood homeomorphic to $\mathbb{R}^2$
(Definition~\ref{def:chartable}).  This is a \emph{local, intrinsic}
property: it depends only on the topology of $X$ near $x$, not on any
embedding, metric, or coordinate system.  The set of
$\mathrm{ChartableR2}$ points is open (Theorem~\ref{thm:chartable-props}a);
its complement---the ``boundary'' in the $\mathrm{ChartableR2}$ sense---is
closed and sequentially stable.

The key observation is that whether a point is $\mathrm{ChartableR2}$ is a
\textbf{finite-data invariant}: it depends on the existence of a single
open neighborhood and a single homeomorphism to $\mathbb{R}^2$.  Two
spaces are either homeomorphic or not; this status does not change with
more computation, more data, or more iterations.  It is a
\emph{topological fact}, not a resource-dependent property.

\paragraph{Why the invariant is impervious to computation.}
Consider a system whose state space has a particular boundary-connectivity
pattern (e.g., two disjoint connected components, like the closed
cylinder).  The following moves \emph{do not change} the topological type
of the state space, and therefore do not change its boundary pattern:

\begin{itemize}[leftmargin=1.5em,itemsep=0.2ex,topsep=0.4ex,parsep=0pt]
\item \textbf{Computational scaling}---more parameters, more layers,
  longer context windows, more training data.  These extend the system
  \emph{within} a fixed topological type; they do not produce a
  homeomorphism to a different type.
\item \textbf{Self-model iteration.}  Building deeper
  meta-levels, stacking self-interpreters, or running
  $\mathrm{represent}^n$ for larger~$n$.  The iterate lives in the
  same endomorphism monoid and does not alter boundary connectivity.
\item \textbf{Architectural reparameterization}---changing activation
  functions, attention patterns, or layer structure.  If the
  reparameterization is continuous (a homeomorphism of the parameter
  space), the topological type is preserved by definition.  If it is
  discontinuous, it is no longer ``the same architecture, tuned
  differently''---it is a genuinely different topological space, which
  is precisely the kind of structural change the barrier demands.
\item \textbf{Model substitution}---choosing a Klein bottle, projective
  plane, or higher-genus non-orientable surface as the target.  The
  general lemma covers all of these: any two spaces with incompatible
  $\mathrm{ChartableR2}$-co-characterized boundaries are not
  homeomorphic, regardless of which specific surfaces are involved.
\end{itemize}

\paragraph{The general lemma as an open-ended class.}
Theorem~\ref{thm:generic} is \emph{not} a statement about the cylinder
and the M\"obius strip.  It is a universally quantified statement: for
\emph{any} topological spaces $X$ and $Y$, if (i)~there exist subsets
$b_X \subseteq X$ and $b_Y \subseteq Y$ that are exactly the
non-$\mathrm{ChartableR2}$ points, and (ii)~the subspace topologies on
$b_X$ and $b_Y$ make them non-homeomorphic, then $X$ and $Y$ are not
homeomorphic.  This defines an \textbf{open-ended class} of obstructed
pairs: the class includes the cylinder/M\"obius pair (the minimal
witness), but it equally includes torus/Klein bottle,
sphere/projective plane, and any pair of surfaces---or higher-dimensional
spaces---that satisfy the three hypotheses.

Any future domain whose state space matches the boundary pattern of one
side of the class inherits the obstruction with no new proof required.
The adversary cannot escape by choosing a different target; the lemma
is waiting for every target that has an incompatible boundary.

\paragraph{``Wrong type,'' not ``inexhaustible queue.''}
The deepest distinction between the topological barrier and
diagonal/regress arguments is this: diagonal arguments say ``you will
always miss something---there is always one more function not in the
range.''  The topological barrier says ``you are the \emph{wrong kind
of thing}---no amount of adding, iterating, or scaling can bridge a
difference in topological type.''  The first is an inexhaustible queue;
the second is a category mismatch.  Both are formally proved, but the
topological barrier is the one that cannot be routed around by ``trying
harder within the same framework.''

\begin{keybox}[title={The topological barrier: wrong type, not inexhaustible queue}]
The general boundary-contrast obstruction
(Theorem~\ref{thm:generic}) defines an \textbf{open-ended class} of
topologically obstructed space-pairs.  The invariant---boundary
connectivity as detected by $\mathrm{ChartableR2}$---is a
\textbf{finite-data topological fact}: no amount of computation,
scaling, iteration, or continuous deformation within a fixed
topological type can alter it.  The cylinder/M\"obius pair is the
\textbf{minimal-case witness}; the argument itself is the
\textbf{general lemma}, which applies to every pair of spaces with
incompatible boundaries---Klein bottles, projective planes,
higher-genus surfaces, and beyond.  The obstacle is not ``you will
always miss one more item'' but ``you are the wrong kind of space.''
\end{keybox}

\subsection{Half-Space Obstructions}
\label{sec:halfspace}

\begin{theorem}[\leanref{HalfPlaneVsPlane.lean}]
\label{thm:halfplane}
\AfterTheoremHead
\TheoremLeanTag{euclideanHalfSpace2\_not\_homeomorphic\_euclidean2}
$\mathrm{IsEmpty}(H^2 \simeq_t \mathbb{R}^2)$.
Proof via punctured half-plane (star-convex $\Rightarrow$ simply connected) vs.\ punctured plane (not simply connected, from $\exp$ covering).
\end{theorem}

\begin{theorem}[\leanref{HalfSpaceNeighborVsPlane.lean}]
\label{thm:halfspace-local}
\AfterTheoremHead
\TheoremLeanTag{isEmpty\_homeomorph\_subtype\_openH2\_zero\_nbhd\_subtype\_openPlane}
For open $V \subseteq H^2$ with $0 \in V$ and open $O \subseteq \mathbb{R}^2$, there is no homeomorphism $V \cong_t O$.
\end{theorem}

\subsection{Cylinder C4: Boundary Biconditional}

\begin{theorem}[\leanref{CylinderChartableBoundary.lean}]
\label{thm:cylinder-c4}
\AfterTheoremHead
\TheoremLeanTag{closedCylinder\_boundaryUnion\_iff\_not\_chartableR2}
For all $x \in \mathrm{ClosedCylinder}$:
\[
  x \in \partial_{\mathrm{Cyl}} \;\iff\; \neg\,\mathrm{ChartableR2}(x).
\]
\end{theorem}

% ============================================================================
% 5. THE SEAM-CHART PROBLEM
% ============================================================================
\section{Interior Charts on the M\"obius Witness: The Seam Problem}
\label{sec:seam}

\subsection{Why the Seam Is Hard}
\label{sec:seam-why}

Interior-height points of the fundamental domain fall into three regimes: strict
interior $0<x<1$, where the quotient is injective on small boxes; the vertical
seam $x\in\{0,1\}$ with $t\neq\tfrac{1}{2}$, where trigonometric coordinates can
separate fibers when the height window avoids the equator; and the seam at
$t=\tfrac{1}{2}$, where $(0,\tfrac{1}{2})\sim(1,\tfrac{1}{2})$ and a trigonometric
chart degenerates. $\mathrm{ChartableR2}$ at the equator quotient class is
obtained by a piecewise Euclidean unfolding map (see below).

\subsection{Open-Cell Charts}

\begin{theorem}[\leanref{MobiusChartableBoundary.lean}]
\label{thm:open-cell}
\AfterTheoremHead
\TheoremLeanTag{chartableR2\_mobiusQuotientMk\_of\_Ioo\_square}
If $p \in (0,1)^2$ (strict interior of the fundamental domain), then $\mathrm{ChartableR2}(\llbracket p \rrbracket)$.
\end{theorem}

\subsection{Seam Trig Charts (Away from Equator)}

\begin{definition}[\leanref{MobiusSeamChart.lean}, \leanref{MobiusSeamChartable.lean}]
The \emph{saturated seam patch} $S(t_0, \varepsilon, \delta)$ is the union of the left and right $\varepsilon$-thickened seam strips at height window $(t_0 - \delta, t_0 + \delta)$, saturated under $\sim$.
\end{definition}

\begin{theorem}[\leanref{MobiusSeamTrigInject.lean}]
\label{thm:trig-inject}
\AfterTheoremHead
\TheoremLeanTag{onPatch\_mobiusFDTrigCoord\_eq\_iff\_mobiusRel0}
When $\delta < |t_0 - \tfrac{1}{2}|$, the trigonometric FD coordinate map $\mathrm{mobiusFDTrigCoord}$ satisfies:\newline
\noindent$\mathrm{mobiusFDTrigCoord}(p) = \mathrm{mobiusFDTrigCoord}(q)$ on the saturated patch if and only if $p \sim q$.
\end{theorem}

\begin{theorem}[\leanref{MobiusSeamChartableR2.lean}]
\label{thm:seam-trig}
\AfterTheoremHead
\TheoremLeanTag{chartableR2\_mobiusQuotientMk\_of\_mem\_mobiusSeamSaturatedPatch}
For $t_0 \neq \tfrac{1}{2}$ and appropriate $\varepsilon, \delta$, every point\newline
\noindent in $\pi(S(t_0, \varepsilon, \delta))$ is $\mathrm{ChartableR2}$.
\end{theorem}

\subsection{The Equator Chart: Piecewise Unfolding}
\label{sec:equator}

\begin{definition}[\leanref{MobiusSeamChartableR2.lean}]
\label{def:unfold}
\leavevmode\par\noindent
The \emph{fundamental-domain unfolding map}\newline
\noindent$\mathrm{unfoldFD} : [0,1]^2 \to \mathbb{R}^2$ is:
\[
  \mathrm{unfoldFD}(x, t) :=
  \begin{cases}
    \mathrm{mobiusPlaneCoord}(x, t) & \text{if } x \leq \tfrac{1}{2}, \\
    \begin{aligned}[t]
      &\mathrm{mobiusPlaneTrigSeamPartnerSub}\\
      &\quad\bigl(\mathrm{mobiusPlaneCoord}(x, t)\bigr)
    \end{aligned} & \text{if } x > \tfrac{1}{2}.
  \end{cases}
\]
\end{definition}

\begin{lemma}[\leanref{MobiusSeamChartableR2.lean}]
$\mathrm{unfoldFD}$ is $\sim$-invariant: if $p \sim q$ then $\mathrm{unfoldFD}(p) = \mathrm{unfoldFD}(q)$.
\end{lemma}

\begin{theorem}[\leanref{MobiusSeamChartableR2.lean}]
\label{thm:equator}
\AfterTheoremHead
\TheoremLeanTag{chartableR2\_mobiusQuotientMk\_of\_left\_seam\_half}
$\mathrm{ChartableR2}(\llbracket (0, \tfrac{1}{2}) \rrbracket)$.
\end{theorem}

\subsection{Assembly: All Interior Points Are ChartableR2}

\begin{theorem}[\leanref{MobiusSeamChartableR2.lean}]
\label{thm:interior-height}
\AfterTheoremHead
\TheoremLeanTag{chartableR2\_mobiusQuotientMk\_of\_interior\_height}
For every $p \in [0,1]^2$ with $0 < p_2 < 1$:
\[
  \mathrm{ChartableR2}(\llbracket p \rrbracket).
\]
\end{theorem}

\subsection{M\"obius C4: Boundary Biconditional}

\begin{theorem}[\leanref{MobiusChartableBoundary.lean}]
\label{thm:mobius-c4}
\AfterTheoremHead
\TheoremLeanTag{mobiusStripBoundary\_iff\_not\_chartableR2\_of\_forall\_off\_edge\_chartable}
For all $x \in \mathrm{MobiusStrip}$:
\[
  x \in \partial_{\mathrm{Mob}} \;\iff\; \neg\,\mathrm{ChartableR2}(x).
\]
\end{theorem}

% ============================================================================
% 6. M-FINAL: THE NON-HOMEOMORPHISM
% ============================================================================
\section{M-FINAL: Minimal-Case Witness for the General Obstruction}
\label{sec:mfinal}

The general lemma (Theorem~\ref{thm:generic}) applies to \emph{any} pair of
spaces with incompatible \texttt{ChartableR2}-co-characterized boundaries.
The M\"obius strip and closed cylinder serve as the \textbf{simplest fully
discharged witness}: a concrete compact pair for which all three hypotheses
are proved.  The argument is not specific to these two surfaces; a Klein
bottle vs.\ torus, a projective plane vs.\ a sphere, or any other pair
meeting the same three conditions would yield the same conclusion.

\begin{theorem}[\leanref{MobiusSeamChartableR2.lean}, \leanref{ChartableR2Bridge.lean}]
\label{thm:mfinal}
\AfterTheoremHead
\TheoremLeanTag{mobiusStrip\_not\_homeomorphic\_closedCylinder}
\[
  \mathrm{IsEmpty}\!\left(\mathrm{MobiusStrip} \simeq_t \mathrm{ClosedCylinder}\right).
\]
\end{theorem}

\begin{proof}[Proof sketch]
Instantiate Theorem~\ref{thm:generic} with $b_X = \partial_{\mathrm{Mob}}$,
$b_Y = \partial_{\mathrm{Cyl}}$, the biconditionals
(Theorems~\ref{thm:cylinder-c4} and~\ref{thm:mobius-c4}), and the boundary
incompatibility (Theorem~\ref{thm:boundary-contrast}).
\end{proof}

% ============================================================================
% 7. THE EXTENDED MASTER THEOREM AND INSTANTIATIONS
% ============================================================================
\section{The Extended Master Theorem and Instantiations}
\label{sec:extended}

The summit theorem (Theorem~\ref{thm:summit}) establishes representational
incompleteness as a pure diagonal result.  This section bundles it with the
topological strand and the horn closures into a single verifiable package,
then draws out the consequences for five neighboring debates.

\subsection{Formal Statement}

\begin{theorem}[\leanref{Main.lean}]
\label{thm:extended}
\AfterTheoremHead
\TheoremLeanTag{representational\_regress\_master\_extended}
For any representational system $R$:
\[
  \mathrm{RepresentationalIncompletenessMasterExtended}(R)
\]
where this abbreviation expands to:
\begin{enumerate}[label=(\roman*)]
\item $\mathrm{representational\_regress\_master\_claim}(R)$\newline
\noindent (regress + wall + $T_2$ transport),
\item $\mathrm{IsEmpty}(H^1 \simeq_t \mathbb{R}^1)$ (1D half-space vs.\ line;\newline
\noindent\leanref{representational\_regress\_topology\_halfLineModel}),
\item $\mathrm{IsEmpty}(\mathrm{MobiusStrip} \simeq_t \mathrm{ClosedCylinder})$ (M-FINAL; the minimal-case witness for the general obstruction class).
\end{enumerate}
\end{theorem}

\noindent
This \emph{extends} \leanref{representational\_regress\_master}; it does not replace it.

\subsection{Bundled boundary models}

\begin{definition}[\leanref{ChartableR2Neighbor.lean}]
A \emph{\texttt{ChartableR2} boundary model} on $X$ is a subset $b_X \subseteq X$
together with a proof that
$\forall x,\; x \in b_X \iff \neg\,\mathrm{ChartableR2}(x)$.
\end{definition}

\noindent
\begin{itemize}[leftmargin=1.75em,itemsep=0.35ex,topsep=0.6ex,parsep=0pt]
\item The type \leanref{ChartableR2BoundaryModel} packages such data; the lemma
  \leanref{ChartableR2BoundaryModel.not\_homeomorphic} restates
  Theorem~\ref{thm:generic} in bundled form.
\item The development \leanref{ChartableR2ConcreteBoundaryModels} supplies\newline
  \leanref{mobiusStripChartableR2BoundaryModel} and
  \leanref{closedCylinderChartableR2BoundaryModel}.
\item The same-file theorem\newline
  \leanref{mobiusStrip\_closedCylinder\_boundary\_models\_not\_homeomorphic}\newline
  re-proves M-FINAL along that pipeline.
\end{itemize}

\subsection{Instantiations}
\label{sec:instantiations}

Theorems earlier in the paper are unconditional. This subsection records how the
same statements bear on six debates outside pure mathematics:
consciousness, symbol grounding, diagonal arguments, quantum measurement,
the simulation hypothesis, and LLM/AI scaling.
Where an extra premise is required, it is stated explicitly; nothing in the
formalization defaults a controversial premise to true.

\medskip
\noindent\textbf{Universal closure note.}
All six horn closures (\S\ref{sec:noescape}) apply to \emph{every}
instantiation below, because the closures constrain the structure of
any parametric self-model $s : A \to A \to B$, not the domain of
application.  The remarks below therefore focus on domain-specific
nuances; the full horn-by-horn analysis is in \S\ref{sec:noescape}.

\subsubsection{Consciousness and self-presence}

\paragraph{Philosophical bridge (premise P2).}
Some accounts distinguish representational self-reference from a stronger notion
of \emph{self-presence} (first-person givenness, non-representational
acquaintance, etc.). Premise~\textbf{P2} is \emph{not} a theorem in the library:
it says only that, for the purposes of a topological comparison, the
representational substrate is modeled by a space whose
\texttt{ChartableR2}-detected boundary has one connectivity pattern (e.g.\
the cylinder's two disjoint components), while the target phenomenon is modeled
by a space whose boundary has a different, incompatible pattern
(Theorem~\ref{thm:generic} applies to any pair with incompatible
boundaries; the M\"obius strip is the minimal witness).

\paragraph{What is proved.}
The engine is Theorem~\ref{thm:generic}, which defines an
open-ended class of obstructed space-pairs (\S\ref{sec:topo-barrier-detail}).
Theorem~\ref{thm:mfinal} is the minimal-case witness; any other
pair satisfying the three hypotheses would serve equally.

\paragraph{Admissibility under P2.}
All six horn closures apply (see universal closure note above and
\S\ref{sec:noescape}).  What remains \textbf{admissible}: the relevant
substrate must \emph{already} have the target topological type, or the
account must invoke a discontinuous or non-topological bridge not modeled
in this formalism.  Scaling, iteration, richer self-models, or
architectural reparameterization within a fixed topological type are
\emph{inadmissible} routes to the target.

\subsubsection{Symbol grounding and semantic regress}

In the symbol-grounding problem, arbitrary symbols must be interpreted, and
interpreting the interpretation threatens a regress unless a non-representational
stop is posited. Formally, a representational system already contains the morphism
\texttt{represent} whose iterates index ``one more representational step.'' The
type alias \leanref{SymbolSystem} is definitionally identical to
\leanref{RepresentationalSystem}; lemmas such as
\leanref{symbolGrounding\_metaRegressArrow\_injective} record that distinct
natural-number depths correspond to \emph{distinct} endomorphisms of the
awareness object.

\paragraph{Admissibility.}
All six horn closures apply (\S\ref{sec:noescape}).  Admissible
grounding must therefore come from \emph{outside} the representational
loop: the system cannot ground itself by iterating its own interpretive
machinery, whether totally or partially, whether with infinite depth or
finite.

\subsubsection{Diagonal arguments and self-interpretation}

Lawvere's theorem in \texttt{Type} implies that there is no family
$s : A \to A \to B$ that realizes every unary $g : A \to B$ as $s(a)$ for some
parameter $a$, once $B$ carries an endomorphism without fixed points.
\leanref{LawvereRegressBridge} collects the instances and connects them to
the categorical presentation;
\leanref{lawvere\_fixed\_point\_stays\_representational} says the
\texttt{OntologicalSlot} wall persists at every iterate depth.

\paragraph{Admissibility.}
All six horn closures apply (\S\ref{sec:noescape}): total
self-interpreters are blocked by Lawvere; partial ones still miss the
diagonal (\leanref{lawvere\_diagonal\_not\_in\_range}); restricting
to $\mathrm{Bool}$ does not help; climbing the iterate tower does
not dissolve the endo/obj wall.  Admissible self-interpreters must
therefore be \emph{incomplete}: they cannot model their own diagonal
behavior, regardless of the codomain, cardinality, or computability
of~$s$.

\subsubsection{Quantum measurement and observer chains}

Talk of apparatus, environment, and Wigner's-friend hierarchies suggests a chain of
systems, each described relative to a larger ``observer.'' If one ever gives such
a chain a \emph{topological} state space $\mathrm{Obs}$, together with a subset
$\partial_{\mathrm{obs}}$ whose membership agrees with
$\neg\,\mathrm{ChartableR2}$, then Theorem~\ref{thm:generic} applies to any
target space whose boundary subspace is not homeomorphic to
$\partial_{\mathrm{obs}}$ (the M\"obius strip is the minimal witness;
see \S\ref{sec:topo-barrier-detail} for the general class).
\leanref{QuantumObserverChainHypothesis} packages exactly this data as
assumptions; the conclusion
$\mathrm{IsEmpty}(\mathrm{Obs} \simeq_t \mathrm{Target})$ follows for
any target satisfying the hypotheses.

\paragraph{Admissibility.}
All six horn closures apply (\S\ref{sec:noescape}).  The chain must
either introduce genuinely new topological structure at some
level---not representational iteration of the same type---or accept
that its boundary organization differs from the target.  The library
does \emph{not} define Hilbert spaces, observables, or projection
postulates; any physical application requires an independent argument
that a concrete model satisfies the hypothesis fields.

\subsubsection{The simulation hypothesis}
\label{sec:simulation}

The simulation hypothesis posits that a sufficiently powerful
computational system could simulate an entire universe, including the
conscious observers within it, to such fidelity that the simulated
beings cannot distinguish simulation from ``base reality.''
Representational incompleteness constrains this hypothesis from two
independent directions.

\paragraph{The diagonal constraint on self-simulating simulators.}
A simulator that simulates \emph{itself}---whether by modeling its own
hardware within the simulation or by running a copy of its own
code---instantiates $s : A \to A \to B$.  Theorem~\ref{thm:summit}
applies: the simulator's self-model necessarily has a diagonal blind
spot.  It cannot predict its own response to queries about its own
diagonal behavior.  This does not prevent \emph{partial} simulation of
the universe (including partial models of the simulator), but it does
prevent a \emph{complete} simulation that includes a faithful copy of
the simulator's own operation.  A fully self-inclusive simulation is
structurally impossible for the same reason a complete self-interpreter
is: the diagonal exclusion is universal over $s$.

\paragraph{The topological constraint on simulated consciousness.}
If conscious self-presence requires a state space with a boundary
pattern incompatible with the simulator's substrate (premise~P2),
then Theorem~\ref{thm:generic} applies: no homeomorphism maps the
simulator's state space to the target, regardless of fidelity or
computational power.  The topological type of the substrate matters;
``sufficiently faithful simulation'' does not override topological
incompatibility.

\paragraph{Closed routes.}
All six horn closures apply (\S\ref{sec:noescape}): running longer
extends within a fixed topological type; nesting simulations produces
iterated $\mathrm{represent}^n$ subject to the Horn~1/2 dichotomy;
offloading to a separate module triggers the section--retraction
closure; partial simulation leaves a diagonal-shaped blind spot.

\paragraph{What is not ruled out.}
A simulation that does not include a complete model of its own
simulator is not constrained by the diagonal exclusion; the simulation
hypothesis in its weak form (``we might be simulated by a system that
does not model itself'') is not addressed.  The topological constraint
applies only if the conscious-experience target has an incompatible
boundary pattern; if one rejects premise~P2, the topological barrier
does not apply to consciousness per se.  The formalism is precise
about what it constrains and what it does not.

\subsection{Scaling of language models and similar systems}

Training larger models extends a representational architecture within a
fixed topological type; it does not produce a homeomorphism to a different
type.  Theorem~\ref{thm:generic} blocks every target with incompatible
boundary (\S\ref{sec:topo-barrier-detail}).

\paragraph{Admissibility.}
All six horn closures apply (\S\ref{sec:noescape}).  Admissible paths
to a topological type with different boundary organization require
\emph{structural} change---a different topology of the underlying state
space---not quantitative extension of the same architecture (more
parameters, deeper layers, longer context windows, finer-grained
self-models).

\subsection{Implications for AI, agents, and AGI}
\label{sec:ai-agi}

The results of this paper bear directly on the most consequential open
question in artificial intelligence: \emph{can a computational system
achieve genuine self-understanding?}  This subsection unpacks the
implications across the spectrum from narrow AI through autonomous agents
to the hypothetical threshold of artificial general intelligence (AGI).

\subsubsection{Bridge hypothesis P3 (computational self-model)}

To apply representational incompleteness to a concrete AI system, one
must exhibit the parametric self-model explicitly.  We formalize this
as \textbf{bridge hypothesis~P3}
(\leanref{ComputationalSelfModel} in \leanref{ClosureProperties.lean}):

\begin{definition}[\novel\ \leanref{ClosureProperties.lean}:
  \leanref{ComputationalSelfModel}]
\label{def:p3}
A \emph{computational self-model} over types $A$, $B$ is a triple
$(s, f, h_f)$ where $s : A \to A \to B$ is the self-model,
$f : B \to B$ is a fixed-point-free endomorphism, and $h_f$ proves
$\forall b,\; f(b) \neq b$.
\end{definition}

\noindent
P3 is not an axiom: it is a \emph{conditional premise} that must be
discharged for each application, just as P2 (the consciousness/topology
bridge) must be discharged for the topological reading.  The formal
consequences are immediate:

\begin{theorem}[\novel\ \leanref{ClosureProperties.lean}]
\label{thm:p3}
\leavevmode
\begin{enumerate}[label=(\alph*)]
\item (\leanref{ComputationalSelfModel.diagonal\_blind\_spot})
  Under~P3, the self-model has a diagonal blind spot.
\item (\leanref{ComputationalSelfModel.no\_complete\_self\_model})
  Under~P3, no surjective self-model exists.
\end{enumerate}
\end{theorem}

\subsubsection{What representational incompleteness says about AI self-models}

Every AI system that constructs a model of itself---whether an
explicit world-model that includes the agent, an internal self-monitor,
a learned representation of its own weights, or a meta-cognitive
module---instantiates P3 with $A$ indexing the system's states or
parameters, $B$ the codomain of the model's predictions, and $s$ the
self-model function.  Theorem~\ref{thm:p3} applies:
the diagonal function $a \mapsto f(s\,a\,a)$ is never any row $s\,a_0$.
The self-model \emph{necessarily} has a blind spot that includes its own
diagonal behavior.

This is not a limitation of current hardware, training data, or
architecture.  It holds for \emph{any} $s$---learned or hand-coded,
neural or symbolic, finite or infinite, computable or not
(\leanref{computability\_independent\_diagonal}).  No amount
of scaling (more parameters, more compute, more data) changes the
diagonal exclusion, because it is a theorem about the function $s$
itself, not about the resources used to compute it.

\subsubsection{Autonomous agents and recursive self-improvement}

Autonomous agents that plan, act, and revise their own strategies face
the self-modeling barrier at every level:

\begin{itemize}[leftmargin=1.5em,itemsep=0.2ex,topsep=0.4ex,parsep=0pt]
\item \textbf{Self-evaluation.}  An agent that evaluates its own
  performance by modeling its own decision function instantiates $s$.
  The diagonal exclusion guarantees a blind spot: the agent cannot
  predict its own response to the question ``what would you predict
  about your own prediction here?''  This is not a practical
  difficulty; it is a structural impossibility.
\item \textbf{Recursive self-improvement.}  An agent that rewrites
  its own code or retrains its own weights produces a \emph{chain}
  of self-models $s_1, s_2, \ldots$  The blind spot persists at
  every link:
  \leanref{recursive\_self\_improvement\_blind\_spot} proves that
  for every chain index $n$ and every $a_0$, the $n$-th model
  misses its own diagonal.  The iterate dichotomy
  (Remark~\ref{rem:dichotomy}) adds: the improvement
  sequence is either injective (infinite regress with the
  Lawvere wall intact) or it cycles (bounded depth,
  no injective tower).
\item \textbf{Multi-agent delegation.}  An agent that delegates
  self-evaluation to a second agent via any delegation map
  $\pi : A_1 \to A_2$ does not escape:
  \leanref{multiagent\_delegation\_blind\_spot} proves the
  delegated model $a_1 \mapsto s_2(\pi(a_1), \pi(\cdot))$ still
  has a diagonal blind spot; the categorical
  section--retraction closure (Horn~3) shows the combined system
  lacks an injective tower on either side.
\end{itemize}

\subsubsection{The AGI threshold}

Claims about artificial general intelligence often center on the
prospect of a system that ``understands itself'' well enough to
improve without limit.  Representational incompleteness constrains
this prospect precisely:

\begin{enumerate}[label=(\alph*),itemsep=0.3ex]
\item \textbf{No complete self-model is possible.}
  Theorem~\ref{thm:summit} proves that for any $s : A \to A \to B$
  with $B$ admitting a fixed-point-free endomorphism, $s$ cannot
  capture its own diagonal.  An AGI that claims to ``fully understand
  itself'' via a parametric self-model simply cannot exist.
\item \textbf{Partial self-models have a diagonal-shaped hole.}
  Horn~4 (\leanref{lawvere\_diagonal\_not\_in\_range}) shows that even
  a partial self-model misses \emph{exactly} the information needed
  for self-capture: the system's own diagonal behavior.  Useful
  partial self-understanding is not ruled out; \emph{complete}
  parametric self-understanding is.
\item \textbf{The topological barrier is orthogonal.}
  Even if the diagonal limitation were somehow circumvented (it
  cannot be, but hypothetically), the topological barrier
  (Theorem~\ref{thm:generic}) stands independently.  If the
  system's state space has one boundary-connectivity pattern and the
  target of ``full self-understanding'' has an incompatible pattern,
  no continuous deformation within the same topological type bridges
  the gap.
\item \textbf{The ontological sort wall persists.}
  The \texttt{OntologicalSlot} wall (Theorem~\ref{thm:wall}) ensures
  that the \emph{process} of self-modeling ($\mathrm{endo}$) and the
  \emph{product} of self-modeling ($\mathrm{obj}$) remain
  definitionally disjoint at every iterate depth.  An AGI that models
  itself is always on the $\mathrm{endo}$ side of the wall; the
  ``self'' it models is always on the $\mathrm{obj}$ side.  The two
  never meet.
\end{enumerate}

\subsubsection{What is \emph{not} ruled out}

The results are precise about what is excluded and what remains open:

\begin{itemize}[leftmargin=1.5em,itemsep=0.2ex,topsep=0.4ex,parsep=0pt]
\item \textbf{Useful partial self-models} are not ruled out.  An AI
  system can have highly accurate, practically valuable models of its
  own behavior---just not \emph{complete} ones.  The blind spot is
  diagonal-shaped, not total.
\item \textbf{Impressive capabilities} are not ruled out.  Nothing
  in the formalism says AI systems cannot be extraordinarily capable
  at tasks that do not require complete self-capture.
\item \textbf{Structural change} remains admissible.  The barrier
  demands structural change to topological type, not quantitative
  extension.  Whether such structural change is physically realizable
  or desirable is an empirical and engineering question, not one
  settled by this formalism.
\end{itemize}

\begin{keybox}[title={Implications for AI and AGI}]
Every AI self-model---whether learned, hand-coded, neural, or
symbolic---is an instance of $s : A \to A \to B$ and is therefore
subject to the diagonal exclusion (Theorem~\ref{thm:summit}): it
necessarily has a blind spot that includes its own diagonal behavior.
Scaling (more parameters, data, compute) does not help.  Recursive
self-improvement either produces an infinite regress with the Lawvere
wall intact or cycles in bounded depth.  Multi-agent delegation does
not escape the section--retraction closure.  The topological barrier
stands independently.  The system's actual diagonal behavior is
generated by its substrate---the computational or physical machinery
executing it---not by its self-model
(\S\ref{sec:substrate-inference}).
Complete parametric self-understanding is structurally impossible;
useful partial self-understanding is not.
\end{keybox}

\subsection{Relationship to the Trans-Computational Processing argument}

An earlier line of work argues from computability---halting,
G\"odel-style diagonalization, and related limits---to constraints on
``self-contained'' computational mind models.  Representational
incompleteness (Theorem~\ref{thm:summit}) is logically
\emph{independent}: it requires only a fixed-point-free endomorphism
on the codomain, not recursion-theoretic machinery.  The topological
strand (Theorem~\ref{thm:generic}) is likewise orthogonal: it turns on
boundary-subspace invariants, not on recursiveness.  The two approaches
can reinforce a common \emph{limitative} conclusion about
representation closed under its own rules, but neither formally entails
the other.  Importantly, the present result is \emph{stronger} in one
respect: it holds without any computability assumption on $s$, whereas
computability-based arguments require $s$ to be computable.

\subsection{Inherited obstructions: computational theories of physics and mind}
\label{sec:inherited}

The preceding instantiations treat the obstructions as constraints on
individual systems---an AI, a simulator, a measurement chain.  This
subsection draws out a broader consequence: any \emph{theory} that
posits a universe containing self-representational systems inherits
the obstructions from within, and is itself subject to them at large.
The compositional inheritance theorems in \leanref{ClosureProperties.lean}
(\leanref{inherited\_diagonal\_exclusion},
\leanref{universe\_model\_inherits\_incompleteness},
\leanref{blind\_spot\_survives\_reindexing},
\leanref{blind\_spot\_survives\_postcomposition},
\leanref{product\_self\_model\_blind\_spot})
provide the formal backbone: the diagonal exclusion propagates through
pre-composition (reindexing), post-composition, embedding, and product
formation.

\subsubsection{Any world-model containing self-modelers inherits the barrier}

Consider a computational theory of physics---digital physics, cellular
automata cosmology, or any framework that models the universe as a
computation $U$.  If $U$ admits \emph{any} subsystem $S \subseteq U$
that models itself (i.e., $S$ instantiates $s : A \to A \to B$),
then Theorem~\ref{thm:summit} applies to $S$ inside $U$: $S$ has a
diagonal blind spot.  But $U$ itself, if it claims to be a faithful
model of the universe that contains $S$, must represent $S$'s behavior
completely---including the diagonal behavior that $S$ itself cannot
represent.  This places a consistency requirement on $U$: either

\begin{enumerate}[label=(\roman*),itemsep=0.2ex]
\item $U$ is \emph{not} itself a parametric self-model (i.e., $U$ does
  not model its own operation via $s_U : A_U \to A_U \to B_U$), in
  which case $U$ is not self-inclusive and the computational theory of
  physics is incomplete in the sense that it cannot model its own
  execution; or
\item $U$ \emph{is} a parametric self-model, in which case
  Theorem~\ref{thm:summit} applies to $U$ directly: $U$ has its own
  diagonal blind spot, and the theory is incomplete in the formal
  sense that it cannot represent its own diagonal behavior.
\end{enumerate}

\noindent
In either case, the obstructions propagate: a universe-model that
contains self-modelers cannot be simultaneously complete, self-inclusive,
and parametric.  The diagonal exclusion is not an artifact of the
subsystem's limitations; it is an inherited structural feature of any
world that hosts self-reference.

The topological barrier (Theorem~\ref{thm:generic}) reinforces this
independently: if $U$'s state space and any subsystem $S$ have
incompatible boundary patterns, no topology-preserving embedding
exists.

\subsubsection{Computational theories of mind and physics: the admissibility filter}

Any theory that posits mind, consciousness, or self-awareness as a
\emph{computational} phenomenon---functionalism, computational
neuroscience, integrated information theory in its computational
variants, global workspace theory with a computational substrate---hits
the same obstructions the moment it admits that the system models itself:

\begin{itemize}[leftmargin=1.5em,itemsep=0.2ex,topsep=0.4ex,parsep=0pt]
\item The diagonal exclusion (Pillar~1) guarantees that the self-model
  has a blind spot.
\item The topological barrier (Pillar~2) guarantees that if the
  substrate has one boundary-connectivity pattern and the target
  phenomenon has an incompatible pattern, no continuous deformation
  within the same topological type bridges the gap.
\item The ontological sort wall (Pillar~3) guarantees that the process
  of self-modeling and the product of self-modeling remain
  definitionally disjoint.
\end{itemize}

A natural reaction is to \emph{weaken} the self-modeling claim: ``the
system does not truly model itself; it only approximates.''  But this
weakening faces a dilemma:

\begin{enumerate}[label=(\alph*),itemsep=0.3ex]
\item \textbf{If the weakened self-model is still parametric}
  ($s : A \to A \to B$ with any $B$ admitting a fixed-point-free
  endomorphism), then Theorem~\ref{thm:summit} applies regardless of
  how ``approximate'' or ``partial'' the model is.  Horn~4
  (\leanref{lawvere\_diagonal\_not\_in\_range}) requires no
  surjectivity: even a partial, approximate self-model misses its own
  diagonal.  The weakening does not escape the obstruction.
\item \textbf{If the weakening removes parametric self-modeling
  entirely}---the theory claims the system has no self-model at
  all---then the theory contradicts the empirical evidence.  Biological
  organisms demonstrably model themselves: proprioception, interoception,
  predictive processing, metacognition, and self-referential thought are
  all well-documented.  A theory of mind that denies self-modeling to
  escape a formal obstruction has not solved the problem; it has
  abandoned the phenomenon it was meant to explain.
\end{enumerate}

\noindent
The same logic applies to computational theories of physics.  If a
theory weakens the role of self-referential subsystems to escape the
obstruction, it must explain why the universe observably contains
systems that model themselves (scientists, instruments, feedback loops).
If it retains self-reference, the obstructions apply.

\subsubsection{The non-computability escape and why topology closes it}
\label{sec:noncomp-escape}

A sophisticated adversary might concede the diagonal exclusion for
computable systems but propose a \emph{non-computable} self-model: ``the
system's self-representation involves a non-computable function, so
G\"odel- and Turing-style arguments do not apply, and perhaps the
diagonal exclusion does not either.''

This escape fails on two levels:

\begin{enumerate}[label=(\arabic*),itemsep=0.3ex]
\item \textbf{The diagonal exclusion is computability-independent.}
  Theorem~\ref{thm:summit} makes no assumption about whether $s$ is
  computable.  The function $s : A \to A \to B$ can be any function
  whatsoever---computable, hyperarithmetic, or wildly non-constructive.
  The diagonal $a \mapsto f(s\,a\,a)$ is still never any row $s\,a_0$.
  Non-computability does not evade Pillar~1.
\item \textbf{The topological barrier is orthogonal to computability.}
  Theorem~\ref{thm:generic} detects a finite-data topological
  invariant---boundary connectivity---that has nothing to do with
  computability or the Church--Turing thesis.  No non-computable
  function can change the homeomorphism type of a space.
\end{enumerate}

\noindent
This is why the topological pillar is especially powerful: it closes
the ``it is just not computable'' escape route that is available
against G\"odel- and Turing-style arguments.  A non-computable oracle
can, in principle, solve the halting problem or prove G\"odel sentences;
it \emph{cannot} change the boundary-connectivity pattern of a
topological space.  The topological barrier is immune to the entire
computability hierarchy.

\begin{keybox}[title={Inherited obstructions and the non-computability escape}]
Any model of a universe that contains self-representational systems
inherits the diagonal exclusion and the topological barrier from
within.  Computational theories of physics and mind hit these
obstructions the moment they admit self-modeling subsystems.
Weakening self-modeling either fails (partial models still have the
diagonal blind spot) or contradicts empirical observation (biological
systems demonstrably model themselves).  Appealing to
non-computability does not help: the diagonal exclusion holds for all
functions regardless of computability, and the topological barrier is
entirely orthogonal to the computability hierarchy.  No non-computable
oracle can change the homeomorphism type of a space.
\end{keybox}

\subsection{What admissible answers look like}
\label{sec:admissible-shape}

The horn closures do not merely exclude; they \emph{shape}.  By
eliminating every quantitative or parametric escape, they delineate
the positive form that any successful account must take.  This
subsection draws out the admissible shape for three central questions.

\subsubsection{Mind, awareness, and self-awareness}
\label{sec:admissible-mind}

If awareness of awareness is modeled as $s : A \to A \to B$ (a system
that maps its own states to representations of its own states), then
the diagonal exclusion guarantees a blind spot: the representation of
``how I represent myself representing myself'' is necessarily absent.
The horn closures block every quantitative remedy.  What remains
admissible?

\begin{enumerate}[label=(\alph*),itemsep=0.3ex]
\item \textbf{Non-representational acquaintance.}
  If self-awareness is not (entirely) a parametric self-model---if
  there is a component of first-person givenness that is not captured
  by \emph{any} $s : A \to A \to B$---then the diagonal exclusion does
  not apply to that component.  This is the route taken by
  phenomenological traditions that distinguish \emph{lived
  self-presence} from \emph{representational self-reference}.  The
  formalism is precise about what it constrains: parametric
  self-models.  Anything that is not a parametric self-model escapes
  by definition---but must then explain what it \emph{is}.
\item \textbf{Structural change to topological type.}
  If the substrate of self-awareness has a topological type that is
  \emph{already} compatible with the target (i.e., the boundary
  pattern matches), then the topological barrier does not apply.  This
  means the substrate was never ``purely representational'' in the
  cylinder sense; it already had the richer structure.
\item \textbf{Acceptance of incompleteness.}
  The most conservative admissible position: self-awareness is a
  genuine parametric self-model, it \emph{does} have a blind spot, and
  that blind spot is constitutive of the phenomenology of
  self-awareness---the ``something it is like'' to be a self-aware
  system includes the structural impossibility of fully capturing
  oneself.  On this reading, representational incompleteness is not a
  defect but a \emph{feature}: the blind spot \emph{is} the gap
  between the knower and the known, formalized.
\end{enumerate}

\noindent
What is \textbf{not} admissible: claiming that self-awareness is a
complete parametric self-model (Theorem~\ref{thm:summit} forbids it);
claiming that enough iteration, scaling, or architectural complexity
will close the gap (Horns~1--6 forbid it); claiming that a
non-computable self-model escapes (\S\ref{sec:noncomp-escape} closes
this).

\begin{keybox}[title={What is mind, given the obstructions?}]
If mind involves self-awareness, and self-awareness involves
self-modeling, then mind is \emph{necessarily incomplete with respect
to itself}.  The blind spot is not a bug to be fixed by better
hardware or deeper recursion; it is a structural feature of any
system that models itself.  Since the system \emph{does} produce
diagonal behavior (it functions), and no self-model row can match
that behavior, the substrate must generate it
(\S\ref{sec:substrate-inference}).  Admissible accounts of mind
must either (a)~include a non-representational component that is
not a parametric self-model, (b)~posit a substrate whose topological
type already matches the target, or (c)~accept the incompleteness as
constitutive---recognizing the blind spot as the formal signature of
the substrate's irreducible role.  Every other route is formally
closed.
\end{keybox}

\subsubsection{The quantum observer}
\label{sec:admissible-qm}

The measurement problem asks: how does a definite outcome emerge when
an observer measures a quantum system?  If the observer is itself a
physical system, it must be describable within the same formalism.
The horn closures constrain what ``describable within the same
formalism'' can mean.

\begin{enumerate}[label=(\alph*),itemsep=0.3ex]
\item \textbf{No complete self-inclusive measurement theory.}
  If the observer's state space admits a parametric self-model
  (the observer models its own measurement process), then
  Theorem~\ref{thm:summit} applies: the observer has a diagonal
  blind spot.  A measurement theory that claims the observer can
  fully predict its own measurement outcomes is structurally
  impossible---for the same reason a complete self-interpreter is.
\item \textbf{Topological constraint on observer--system
  identification.}
  If the observer's state space has one boundary-connectivity pattern
  and the measured system's state space has an incompatible pattern,
  then no homeomorphism identifies them
  (Theorem~\ref{thm:generic}).  This constrains any interpretation
  that treats observer and system as ``the same kind of thing'':
  they can be the same kind only if their topological types are
  compatible.
\item \textbf{Admissible resolutions.}
  The Wigner's-friend hierarchy (stacking observers) is an instance
  of iterated $\mathrm{represent}^n$, subject to the Horn~1/2
  dichotomy.  The admissible resolutions must either
  (i)~introduce genuinely new topological structure at some level
  (not iteration of the same type),
  (ii)~accept that the observer's description of the measurement is
  necessarily incomplete (diagonal blind spot), or
  (iii)~deny that the observer models its own measurement process
  parametrically (exiting the scope of P3).
\end{enumerate}

\noindent
The formalism does not adjudicate between interpretations of quantum
mechanics (Copenhagen, many-worlds, relational, etc.).  It constrains
the \emph{structural form} of any interpretation that includes
self-measuring observers: such an interpretation must accept
representational incompleteness or deny that the observer's
self-measurement is parametric.

\subsection{The anatomy of the blind spot}
\label{sec:blind-spot-anatomy}

The preceding subsections establish what the blind spot \emph{excludes}
and what routes \emph{to} it are closed.  This subsection asks the
harder question: what \emph{is} inside the blind spot, and what does
its existence imply about any system that functions despite having one?

\subsubsection{What is inside the blind spot}

The blind spot has a precise shape.  For any self-model
$s : A \to A \to B$ and any fixed-point-free $f : B \to B$, the
excluded function is:
\[
  g_f : A \to B, \qquad g_f(a) := f(s\,a\,a).
\]
This is the system's own \textbf{diagonal behavior, twisted by $f$}.
In concrete terms:
\begin{itemize}[leftmargin=1.5em,itemsep=0.2ex,topsep=0.4ex,parsep=0pt]
\item If $B = \mathbb{N}$ and $f = \mathrm{succ}$, then $g_f(a)$ is
  ``what I predict about my own state~$a$, \emph{plus one}.''  The
  system cannot have a row of its self-model that matches its own
  prediction-about-its-own-prediction, shifted by one.
\item If $B = \mathrm{Bool}$ and $f = \lnot$, then $g_f(a)$ is
  ``the negation of my self-classification at~$a$.''  No row of
  the self-model matches the system's own self-classification, inverted.
\item In general, the blind spot contains every function that
  \emph{depends on the self-model's own diagonal entries and then
  departs from them}.  The self-model can approximate arbitrarily
  many off-diagonal rows, but the diagonal row---the one where the
  system evaluates itself evaluating itself---is precisely where the
  model fails.
\end{itemize}

\noindent
The blind spot is therefore not random or diffuse.  It is
\textbf{structurally located at the self-referential fixed point}: the
exact place where the system's model turns back on itself.  The
information the system cannot represent is precisely the information
about \emph{how it represents itself}.

\subsubsection{Can a system function with a blind spot?}
\label{sec:functioning}

A self-aware system evidently \emph{does} function.  Biological
organisms exhibit proprioception, metacognition, and recursive
self-referential thought.  AI systems build and use self-models.
The blind spot does not prevent functioning; it constrains the
\emph{fidelity} of self-modeling at the diagonal.

Here is the key formal observation.  The system has a self-model
$s$.  When confronted with a self-referential input---an input in
the diagonal region---the system must produce \emph{some} output.
Call this actual output $b_{\mathrm{actual}}(a)$.  Theorem~\ref{thm:summit}
guarantees:
\[
  b_{\mathrm{actual}} \neq s\,a_0 \quad\text{for every } a_0 \in A.
\]
That is: the system's actual diagonal behavior is \textbf{never
faithfully represented by any row of its own self-model}.  The system
\emph{acts} on diagonal inputs, but it cannot \emph{model} its own
action on those inputs.  There is no contradiction with functioning:
the system functions, but its account of its own functioning is
necessarily inaccurate at the diagonal.

\subsubsection{The substrate inference}
\label{sec:substrate-inference}

The preceding observation leads to a consequence that is both simple
and far-reaching.

\begin{enumerate}[label=(\arabic*)]
\item The system produces diagonal behavior $b_{\mathrm{actual}}$.
\item No row of the self-model $s$ matches $b_{\mathrm{actual}}$.
\item Therefore, the process that \emph{generates} the system's
  diagonal behavior is not captured by $s$.
\end{enumerate}

\noindent
Since $s$ is the system's self-model---the system's own representation
of its own behavior---the process generating the diagonal behavior
must be \textbf{outside the scope of the self-model}.  It is produced
by the system's \emph{substrate}: the underlying machinery
(biological, computational, physical) that executes the system's
behavior, including the behavior the self-model cannot represent.

This is not a metaphysical claim.  It is a direct consequence of
the theorem.  If a system has a parametric self-model $s$ and
produces diagonal behavior, then there exists a component of its
behavior that is generated by substrate-level processes not
representable in $s$.  The self-model is a map; the diagonal
behavior is territory that the map cannot chart.

\paragraph{Formal statement of the substrate inference.}
Let $s : A \to A \to B$ be any self-model, $f : B \to B$
fixed-point-free.  Define the \emph{diagonal gap function}
$\delta_f : A \to B$ by $\delta_f(a) := f(s\,a\,a)$.  Then:
\begin{enumerate}[label=(\alph*),itemsep=0.2ex]
\item $\delta_f \notin \mathrm{range}(s)$
  (Theorem~\ref{thm:summit}(b)).
\item If the system's actual behavior on diagonal inputs is
  $\delta_f$, then that behavior is \emph{not the output of any
  self-model row}---it is generated by processes outside $s$.
\item This holds for \emph{every} $s$, \emph{every} $f$, and
  \emph{every} $A$ and $B$ satisfying the hypotheses.  No change
  of self-model eliminates the gap; it merely moves it
  (\leanref{recursive\_self\_improvement\_blind\_spot}).
\end{enumerate}

\noindent
Point~(b) is where the formal and the empirical meet.  The theorem
provides (a) unconditionally.  The application of (b) requires the
empirical premise that the system \emph{does} exhibit diagonal
behavior---that it processes self-referential inputs and produces
outputs.  Given that premise, (b) follows deductively: the substrate
is doing work that the self-model cannot represent.

\subsubsection{Second-order consequences}

The substrate inference has several downstream implications.

\paragraph{1.\ Self-awareness as representation is not fully self-transparent.}
If self-awareness operates \emph{as a parametric self-model}---if
the system represents itself to itself via some $s : A \to A \to B$---then
it is necessarily opaque to itself at the diagonal.  The ``feeling
of understanding oneself'' cannot be a complete readout of $s$,
because $s$ is missing exactly the self-referential behavior.

There is, however, a precise escape.  The theorem's hypothesis is
``the system has a parametric self-model $s$.''  If self-awareness
is \textbf{not} a parametric self-model but rather the ontological
ground itself---the substrate activity, the \emph{process} of
awareness rather than a \emph{representation} of that process---then
the hypothesis is not satisfied and the blind spot does not apply.
This is exactly the distinction formalized by \texttt{OntologicalSlot}
(Pillar~3): the \emph{endo} side (process) and the \emph{obj} side
(product/representation) are definitionally disjoint
(Theorem~\ref{thm:wall}).  Self-awareness as the ontological ground
would live on the \emph{endo} side of the wall.  It escapes the
blind spot \emph{precisely because} it is not a representation---it
is the territory, not the map.

The formal landscape therefore yields a sharp dichotomy:
\begin{itemize}[leftmargin=1.5em,itemsep=0.15ex,topsep=0.3ex,parsep=0pt]
\item \textbf{Self-awareness as representation} ($s : A \to A \to B$):
  has a diagonal blind spot, is necessarily partial, and requires the
  substrate to generate the diagonal behavior it cannot represent.
\item \textbf{Self-awareness as ontological ground} (the substrate
  activity itself): is not a parametric self-model, is not subject to
  the diagonal exclusion, and lives on the \emph{endo} side of the
  \texttt{OntologicalSlot} wall---definitionally disjoint from any
  representational product.
\end{itemize}
This dichotomy is not a limitation of the formalism.  It \emph{is}
the formalism's central structural prediction: the process of
self-awareness and the representation of self-awareness inhabit
formally disjoint ontological categories, and only the
representational side is subject to the blind spot.

\paragraph{2.\ The gap is generative, not merely privative.}
The blind spot is not a mere absence of information.  It is the
locus of a \textbf{positive mismatch}: the system's actual
behavior at the diagonal \emph{differs from} what any self-model
row predicts.  This mismatch is the formal analogue of the
phenomenological intuition that self-awareness involves a sense of
\emph{surplus}---a feeling that the self exceeds its own
representation.  The formalism makes the surplus precise: it is
exactly the diagonal gap function $\delta_f$.

\paragraph{3.\ No self-improvement program closes the gap.}
An agent that builds successively better self-models $s_1, s_2,
\ldots$ does not converge on the diagonal.  Each $s_n$ has its own
diagonal blind spot
(\leanref{recursive\_self\_improvement\_blind\_spot}).  The gap
does not shrink; it \emph{moves}.  Each new self-model captures
what the previous one missed, but introduces a new diagonal that
it itself cannot capture.  This is the formal backbone of the
intuition that self-understanding is an asymptotic process: one
can always understand more of oneself, but never all.

\paragraph{4.\ Multi-agent pooling does not close the gap.}
An adversary might propose: ``Agent~1 has a blind spot; Agent~2
models Agent~1's blind spot.''  But Agent~2's model of Agent~1 is
itself a parametric model, and
\leanref{multiagent\_delegation\_blind\_spot} proves it has its
own diagonal blind spot.  The combined system of agents has
strictly more modeling capacity, but \emph{not} complete
self-coverage: the pooled diagonal is still unrepresented.

\paragraph{5.\ Something must fill the gap---and it is the substrate.}
If a self-aware system functions (and it evidently does), then
diagonal behavior is produced.  Since no self-model row produces
it, the substrate does.  This is not an optional philosophical
gloss; it is a deductive consequence of the theorem plus the
empirical premise that the system functions.  The substrate---the
physical, biological, or computational machinery underlying the
system---is the \textbf{necessary complement} to the self-model.
Without it, the system cannot produce its own diagonal behavior.

\begin{keybox}[title={What fills the blind spot}]
The blind spot is located at the self-referential diagonal: the
point where the system's model turns back on itself.  The system
\emph{does} produce diagonal behavior (it functions), but its
self-model \emph{cannot} represent that behavior
(Theorem~\ref{thm:summit}).  Therefore, the system's diagonal
behavior is generated by substrate-level processes outside the
scope of any self-model.  The substrate is not optional: it is the
\textbf{formally necessary complement} to any parametric
self-model.  No self-improvement, multi-agent delegation, or
scaling can substitute for it.  The blind spot is not a bug; it is
the \emph{signature} of the substrate's irreducible contribution
to the system's behavior.
\end{keybox}

\subsubsection{Open questions}
\label{sec:open-questions-blind-spot}

Several questions raised by the substrate inference cannot be settled
within this formalism.  We state them explicitly, with the formal
boundary between what is proved and what remains open.

\begin{enumerate}[label=(\roman*),itemsep=0.3ex]
\item \textbf{What is the substrate?}  The formalism is agnostic
  about whether the substrate is physical matter, a computational
  layer below the self-model, or something else entirely.  It proves
  only that \emph{some} process outside the self-model generates the
  diagonal behavior.  Identifying that process is an empirical
  question.
\item \textbf{Is the substrate itself modelable?}  A natural follow-up:
  can the system (or another system) build a model of the substrate?
  If so, that model is itself a parametric function and
  has its own diagonal blind spot
  (\leanref{inherited\_diagonal\_exclusion}).  The substrate, once
  modeled, generates a new diagonal gap.  This suggests an infinite
  regress of substrates---formally captured by the iterate
  dichotomy (Horns~1/2).
\item \textbf{Does the blind spot have phenomenological content?}
  Some phenomenological traditions identify a ``pre-reflective
  self-awareness'' that is not a representational act but a
  background condition of consciousness.  The formalism is
  compatible with this: the blind spot could be the formal shadow
  of the pre-reflective layer.  But the formalism cannot prove this
  identification; it can only prove that \emph{something} outside
  the self-model is necessary.
\item \textbf{Is awareness of awareness possible?}  The formalism
  does not say awareness of awareness is impossible.  It constrains
  only the \emph{representational} mode: awareness of awareness
  \emph{qua parametric self-model} has a diagonal blind spot.  But
  if awareness of awareness is the ontological ground's own
  reflexive activity---living on the \emph{endo} side of
  \texttt{OntologicalSlot}---then it is not a parametric self-model
  and the exclusion does not apply (\S\ref{sec:blind-spot-anatomy},
  consequence~1).  Empirically, awareness of awareness \emph{does}
  occur.  The formalism therefore predicts a disjunction: either
  (a)~it occurs with a diagonal blind spot (the representational
  reading), or (b)~it occurs as ontological-ground activity that
  is not subject to the exclusion (the non-representational reading).
  On either reading, the evidence is consistent.  What the formalism
  \emph{rules out} is a single parametric self-model that is fully
  transparent to itself---perfectly self-aware with no structural
  opacity.  That is the only thing that is formally impossible.
\end{enumerate}

\subsection{Why This Is Not G\"odel Repackaged}
\label{sec:godel-distinction}

An adversarial reader's first instinct on encountering a
``self-reference impossibility'' result may be: \emph{``This is just
G\"odel incompleteness in different notation---nothing new.''}  This
subsection walks through the comparison in detail and explains exactly
what G\"odel alone does \emph{not} give you.

\subsubsection{The structural comparison}

\begin{center}
\setlength{\tabcolsep}{4pt}
\begin{tabular}{@{}>{\raggedright\arraybackslash}p{6.5cm}
                 >{\raggedright\arraybackslash}p{7.5cm}@{}}
\toprule
\textbf{G\"odel Incompleteness} & \textbf{Representational Incompleteness} \\
\midrule
Applies to formal theories with provability predicates &
Applies to \emph{any} parametric self-model $s : A \to A \to B$ \\[0.5ex]
Requires arithmetic ($\Sigma_1$-completeness) &
Requires only a fixed-point-free endomorphism on $B$ \\[0.5ex]
Self-referential \emph{sentence} (the G\"odel sentence $G$) &
Self-referential \emph{function} (the diagonal $a \mapsto f(s\,a\,a)$) \\[0.5ex]
Blind spot is an \emph{unprovable truth} &
Blind spot is an \emph{unrepresentable behavior} \\[0.5ex]
Escape route: extend the theory (add axioms, move to a stronger system) &
No escape: diagonal exclusion holds for \emph{every} $s$, every $A$, partial or total \\[0.5ex]
Computability-dependent (requires recursive axiomatizability or computable functions) &
Computability-\emph{independent} (holds for all functions, computable or not) \\[0.5ex]
Cardinality-dependent (requires countable language) &
Cardinality-independent (holds for every cardinality of $A$) \\
\bottomrule
\end{tabular}
\end{center}

\subsubsection{Why G\"odel's incompleteness is insufficient for our claim}

G\"odel's first incompleteness theorem says: any consistent, recursively
axiomatizable theory $T$ extending Robinson arithmetic contains a sentence
$G$ that is true in the standard model but not provable in $T$.  The
standard remedy is to \emph{extend the theory}: add $G$ as an axiom,
obtaining $T' = T + G$.  Of course $T'$ has its own unprovable sentence
$G'$, but the point is that for any \emph{particular} sentence, one can
always find a stronger theory that proves it.

Representational incompleteness has no such escape.
Theorem~\ref{thm:summit} says: for \emph{any} $s : A \to A \to B$ and any
fixed-point-free $f$, the diagonal $a \mapsto f(s\,a\,a)$ is not $s\,a_0$
for \emph{any} $a_0$.  Making $s$ ``stronger'' (total, surjective, defined
on a larger $A$) does not help; if $s$ is claimed to be surjective, it
simply cannot exist.  There is no analogue of ``add an axiom'': the
diagonal exclusion is universal over $s$.

\subsubsection{Why the Lawvere connection is classical but the packaging is new}

Lawvere's fixed-point theorem (1969) is the categorical ancestor of
Cantor's theorem and, by extension, of G\"odel's diagonalization.  We
freely acknowledge this heritage.  What is new is:

\begin{enumerate}[label=(\alph*),itemsep=0.3ex]
\item The \textbf{\texttt{OntologicalSlot}} construction
  (Definition~\ref{def:ontoslot}), which tags morphisms and global
  elements as belonging to disjoint ontological sorts.  Lawvere's theorem
  does not tag the fixed point; standard category theory does not
  introduce a formal disjointness marker between $\mathrm{End}(A)$ and
  $\mathrm{Hom}(\mathbf{1}, A)$.  See
  \S\ref{sec:ontoslot-novelty}.
\item The \textbf{partial-model diagonal exclusion}
  (\leanref{lawvere\_diagonal\_not\_in\_range}): the blind spot persists
  even when $s$ is not claimed to be surjective.  Classical Lawvere
  requires surjectivity as a hypothesis; our result drops it entirely.
\item The \textbf{decidable / $\mathrm{Bool}$-valued horn closure}
  (Horn~6): even restricting to two-valued predicates---the simplest
  possible codomain---does not help.
  $\lnot : \mathrm{Bool} \to \mathrm{Bool}$ is fixed-point-free, so
  the obstruction persists in the finite, decidable setting.  This
  further separates the result from the G\"odel/Turing tradition, where
  undecidability is essential.
\end{enumerate}

\subsubsection{Computability independence}

G\"odel's theorem requires the system under study to be a recursively
axiomatizable formal theory (or, in Turing's version, a computable
function).  If the system is not recursively enumerable, the theorem does
not apply.

Representational incompleteness makes \emph{no computability assumption}
on $s$.  The function $s : A \to A \to B$ can be any function
whatsoever---computable, hyperarithmetic, $\Delta^1_1$, or wildly
non-constructive.  The diagonal exclusion holds regardless.  In this
precise sense, the result is \emph{stronger} than G\"odel's: it covers
systems that G\"odel's theorem cannot even address.

\subsubsection{The topological strand has no G\"odel analogue at all}

The topological barrier (Theorem~\ref{thm:generic}) is completely
independent of diagonalization.  G\"odel, Cantor, Lawvere---none of them
say anything about homeomorphism types or boundary-connectivity
invariants.  The topological strand detects an obstruction that is
invisible to any proof-theoretic or recursive-function-theoretic argument.
It is a finite-data invariant of a different mathematical universe entirely.

\begin{keybox}[title={Not G\"odel repackaged: a summary}]
Representational incompleteness differs from G\"odel incompleteness
in \textbf{six independent respects}: (1)~no formal theory required,
(2)~no arithmetic required, (3)~no coding required, (4)~the blind
spot is a function not a sentence, (5)~no escape via theory
extension, (6)~no computability assumption.  The topological barrier
adds a \textbf{seventh}: it is orthogonal to all diagonal arguments
and detects a finite-data invariant that no amount of computation,
scaling, or iteration can alter.  Any claim that ``this is just
G\"odel'' must address all seven distinctions simultaneously.
\end{keybox}

\subsection{No-escape analysis: horn-by-horn closure}
\label{sec:noescape}

An adversary confronting the representational regress has, at first glance,
several avenues of retreat.  Each is formalized and shown to fail.  The
numbering below follows the order in which the escape was identified; each
``horn'' names an adversarial move, and the closing theorem blocks it.

\paragraph{Horn~1: Infinite injective tower.}
\emph{Move:} accept $\mathrm{iter\_injective}$ and try to use the tower itself
as a self-representation resource.
\emph{Closure:} Theorem~\ref{thm:regress} yields infinitely many distinct
meta-levels;
Theorem~\ref{thm:wall} (the Lawvere wall) ensures the endo/obj distinction
persists at every level.  The tower does not collapse into closure.

\paragraph{Horn~2: Finite-order / periodic iterates.}
\emph{Move:} weaken $\mathrm{iter\_injective}$---let
$\mathrm{represent}^n = \mathrm{represent}^m$ for some $n \neq m$, so the
loop has only finitely many distinct powers.\newline
\noindent\emph{Closure:}
\leanref{IterateClosure.End.pow\_iterate\_dichotomy}
proves the dichotomy: either the iterates are all distinct (Horn~1) or
there exists a bound $B$ such that every power equals some $e^i$ with
$i < B$.  In the finite case the system's representational depth is
bounded; it cannot host an injective tower at all.

\paragraph{Horn~3: Cross-object readout via section--retraction.}
\emph{Move:} represent $A$ \emph{inside} a different object $B$ using a
section $s : A \to B$ and retraction $r : B \to A$ with $s \circ r = \mathrm{id}_A$.
\emph{Closure:}
\leanref{CrossObjectRepresentation.section\_retraction\_no\_injective\_tower\_either\_side}
shows that $r \circ s$ is idempotent on $B$ (so $(r \circ s)^1 = (r \circ s)^2$,
blocking injective powers on the host side) and $s \circ r = \mathrm{id}_A$
(so all powers are the identity on the source side---trivially periodic).
Neither side supports an injective $\mathbb{N}$-tower.

\paragraph{Horn~4: Partial (non-surjective) self-model.}
\emph{Move:} concede that no \emph{total} self-interpreter exists, but claim
a useful \emph{partial} one.
\emph{Closure:}
\leanref{lawvere\_diagonal\_not\_in\_range}
requires no surjectivity of $s$: for \emph{any} $s : A \to A \to B$ and any
fixed-point-free $f : B \to B$, the diagonal function $a \mapsto f(s\,a\,a)$
is never in the range of $s$.  Every partial self-model necessarily has a
blind spot that includes its own diagonal---the information that matters most
for self-capture is exactly the information it cannot represent.

\paragraph{Horn~5: Different topological models.}
\emph{Move:} reject the cylinder/M\"obius witness; substitute a
Klein bottle, projective plane, or another geometric model.

\emph{Closure:} Theorem~\ref{thm:generic}\newline
(\leanref{not\_homeomorphic\_of\_chartableR2\_boundary\_contrast})\newline
\noindent is \emph{model-agnostic}: it defines an
\textbf{open-ended class} of obstructed pairs---any two spaces
whose \leanref{ChartableR2}-co-characterized boundaries are
subtype-incompatible.  The cylinder/M\"obius pair is the
\emph{minimal-case witness}, not the argument itself.  Replacing
one non-orientable surface with another does not evade the
lemma; it merely changes which boundary incompatibility is
invoked.

\paragraph{Horn~6: Computability / decidable self-reference.}
\emph{Move:} restrict to decidable predicates or finite-valued functions,
where ``self-reference'' seems more tractable than Cantorian diagonals.
\emph{Closure:}
\leanref{no\_universal\_parametric\_unary\_bool}
shows that no $s : A \to A \to \mathrm{Bool}$ can realize every
$g : A \to \mathrm{Bool}$, since $\lnot : \mathrm{Bool} \to \mathrm{Bool}$
is fixed-point-free.  The obstruction is not an artifact of infinite
cardinality; it holds in the decidable, two-valued setting.

\paragraph{Residual openings (not closed here).}
Two further routes are not formally addressed in this paper:
\emph{(i)}~\emph{higher-categorical self-maps} (a self-representation
living in a $2$-category or $\infty$-category, where the monoid of
endomorphisms is replaced by a richer hom-object); and
\emph{(ii)}~\emph{non-categorical frameworks} (a system that resists
categorical encoding entirely).  The first is subject to analogous
monoid-element arguments once a truncation to the $1$-categorical
endomorphism monoid is performed; the second falls outside the scope of
any categorical formalization by definition.  Both are noted as
limitations in \S\ref{sec:limits}.

\subsection{Admissibility: what the closures tell neighboring debates}
\label{sec:admissibility}

Given that Horns~1--6 are closed (\S\ref{sec:noescape}), the formal
landscape constrains what \emph{admissible} solutions look like.  The
reasoning is the same in every case: the closures establish a structural
barrier, so any proposed solution must either (a)~operate outside the
barrier's hypotheses, or (b)~accept the consequences.  The admissible
shape is domain-specific; the inadmissible moves are universal.

\paragraph{Common inadmissible moves (all debates).}
Scaling, iterating, partial modeling, cross-object readout,
finite-depth loops, decidable restriction, and model substitution are
all formally blocked (Horns~1--6).  Non-computability does not help:
the diagonal exclusion is computability-independent, and the
topological barrier is orthogonal to the entire computability hierarchy
(\S\ref{sec:noncomp-escape}).

\paragraph{Admissible shapes (per debate).}
\begin{itemize}[leftmargin=1.5em,itemsep=0.2ex,topsep=0.4ex,parsep=0pt]
\item \textbf{Consciousness / self-presence:} the substrate must
  \emph{already} have the target topological type, or a
  non-topological bridge must be invoked.
\item \textbf{Symbol grounding:} grounding must come from
  \emph{outside} the representational loop.
\item \textbf{Self-interpretation:} admissible self-interpreters
  \emph{must} be incomplete---they cannot model their own diagonal.
\item \textbf{Quantum measurement:} the chain must introduce genuinely
  new topological structure or accept incomplete self-description.
\item \textbf{Simulation hypothesis:} admissible accounts must
  exclude the simulator from its own simulation or accept the blind spot.
\item \textbf{LLM / AI scaling:} admissible paths require
  \emph{structural} change to the state-space topology, not
  quantitative extension.
\item \textbf{AGI self-understanding:} complete parametric
  self-understanding is impossible; useful partial self-understanding
  remains possible (see \S\ref{sec:ai-agi}).
\item \textbf{Computational theories of physics and mind:} any
  theory admitting self-modeling subsystems inherits the obstructions
  (\S\ref{sec:inherited}); weakening self-modeling either fails
  (partial models still miss the diagonal) or contradicts empirical
  evidence.
\end{itemize}

% ============================================================================
% 8. LIMITATIONS AND SCOPE
% ============================================================================
\section{Limitations and scope}
\label{sec:limits}

\begin{enumerate}
\item \textbf{Hypothesis, not derivation.} $\mathrm{iter\_injective}$ is bundled,
  not proved from minimal data. Not every $f : A \to A$ in every category satisfies it.
  The regress is conditional on this non-triviality hypothesis.

\item \textbf{Lawvere in \texttt{Type}, not abstract CCC.} The formalization uses
  $\mathrm{Type}\ u$ and\newline
  $\mathrm{MonoidalClosed}(\mathrm{Type}\ u)$, not a general closed category over arbitrary~$C$.
  A fully abstract CCC version remains future work.

\item \textbf{\texttt{OntologicalSlot} is definitional, not metaphysical.} The disjointness $\mathrm{endo} \neq \mathrm{obj}$ is true by constructor discrimination. The robustness theorems (\leanref{ClosureProperties.lean}) prove this is a structural invariant---it survives any injective relabeling and only collapses under a non-injective map---but the underlying mechanism is type-theoretic, not a deep mathematical theorem.

\item \textbf{Models are chosen, not unique.} The closed cylinder and M\"obius strip are the \emph{minimal-case witnesses} chosen for clarity and tractability.  Theorem~\ref{thm:generic} defines an open-ended class of obstructed pairs; any two spaces with incompatible \texttt{ChartableR2}-co-characterized boundaries (Klein bottle vs.\ torus, projective plane vs.\ sphere, etc.) would serve equally.  Discharging additional witnesses requires proving their boundary biconditionals and subtype incompatibility.

\item \textbf{P2 is philosophical.} The premise ``consciousness requires non-orientable topology'' is not a theorem. Section~\ref{sec:instantiations} states only what follows when P2 is adjoined to the proved non-homeomorphism; accepting or rejecting P2 remains an independent issue.

\item \textbf{No $\mathrm{IsManifold}$ polish.} The M\"obius strip is not given a full Mathlib $\mathrm{IsManifold}$ instance. The proof uses $\mathrm{ChartableR2}$ (open neighborhoods homeomorphic to $\mathbb{R}^2$) as a self-contained predicate, sidestepping the need for a full manifold-with-boundary atlas. This is sufficient for the non-homeomorphism but less elegant than a full atlas approach.

\item \textbf{\texttt{QuantumObserverChainHypothesis}.} This definition only
  bundles topological assumptions (boundary set, \texttt{ChartableR2}
  biconditional,\newline subtype non-homeomorphism to an incompatible target
  boundary---the M\"obius boundary is the minimal witness).  It does not
  import Hilbert spaces, operators, or measurement axioms; a physical
  application would need a separate argument that a concrete model satisfies
  those fields.

\item \textbf{Partial self-models: blind spot, not total impossibility.}
  \leanref{lawvere\_diagonal\_not\_in\_range} proves that every partial
  self-model misses its own diagonal; it does not prove that partial
  self-models are \emph{useless}---only that they are necessarily incomplete
  in a precise, diagonal-shaped way.  Whether that incompleteness matters
  for a given application is a separate question.

\item \textbf{Higher-categorical and non-categorical frameworks.}
  The horn closures rely on $1$-categorical endomorphism monoids.  A
  self-representation living in a $2$-category or $\infty$-category
  replaces the monoid with a richer hom-object; the present arguments
  apply after truncation to the underlying $1$-categorical monoid, but
  a native higher-categorical treatment is not given.  Systems that
  resist categorical encoding entirely fall outside any categorical
  formalization by definition.

\item \textbf{Computability: decidable analogue only.}
  \leanref{no\_universal\_parametric\_unary\_bool} closes the
  $\mathrm{Bool}$-valued Lawvere horn.  A full treatment of Turing-degree
  obstructions (Rice's theorem, Kleene's recursion theorem) in the same
  framework is not attempted; the present result is the finite/decidable
  endpoint of the diagonal family.

\item \textbf{Classical logic.}  The iterate dichotomy (Horn~2) and
  several auxiliary results use classical reasoning (law of excluded
  middle), which is standard in Lean~4's axiom environment.  The
  diagonal exclusion itself
  (\leanref{diagonal\_exclusion\_is\_constructive} in
  \leanref{ClosureProperties.lean}) is constructive: it requires only
  function application and the fixed-point-freeness hypothesis.

\item \textbf{P3 (computational self-model) is a modeling claim.}
  Bridge hypothesis~P3 (Definition~\ref{def:p3}) makes explicit
  the modeling assumption ``the system has a parametric self-model
  $s : A \to A \to B$.''  Whether a given AI system, brain, or
  simulator actually satisfies P3 is an empirical and
  philosophical question, not one settled by this formalism.

\item \textbf{Pillars 1 and 2 are formally independent.}  The
  diagonal exclusion (Pillar~1) and the topological barrier (Pillar~2)
  are proved as separate theorems that do not reference each other.
  The \leanref{TopologicalSelfModelingSystem} structure
  (\leanref{ClosureProperties.lean}) formally conjoins both
  obstructions---when a self-modeling system's state space is not
  homeomorphic to a target, both the diagonal blind spot and the
  non-homeomorphism hold simultaneously---but neither pillar
  \emph{implies} the other.  They constrain the same informal target
  from orthogonal mathematical directions.
\end{enumerate}

% ============================================================================
% 9. RELATED WORK
% ============================================================================
\section{Related work}
\label{sec:related}

\subsection{Formalized topology}
Mathlib's manifold and topology libraries underpin the half-space and chart
lemmas; other proof assistants also host substantial topology.  The weight
here is on a \emph{general} non-homeomorphism lemma (Theorem~\ref{thm:generic})
that defines an open-ended obstruction class, together with a fully discharged
minimal-case witness, rather than on novel classification of differential
invariants.

\subsection{Lawvere's theorem in proof theory}
Lawvere's fixed-point argument appears in multiple formal developments.
\texttt{OntologicalSlot} and the wall lemmas supply complementary
category-internal tagging alongside the \texttt{Type}-level diagonal
theorem.  Crucially, Lawvere's original result does not tag the fixed
point as belonging to a different ontological sort from the morphism
that produced it; \texttt{OntologicalSlot}
(Definition~\ref{def:ontoslot}) fills this gap with an inductive
disjointness marker that is, to our knowledge, absent from all prior
formalizations of Lawvere's theorem
(see \S\ref{sec:ontoslot-novelty}).

\subsection{Consciousness, G\"odel, and AI}
G\"odel- and computability-based limits on mechanism (including Penrose-style
arguments, discussions of strange loops and self-reference in the sense of
Hofstadter, and information-theoretic theories of
consciousness) occupy a large literature.  Those arguments turn on
decidability or information capacity; ours turns on topological type.
Representational incompleteness (Theorem~\ref{thm:summit}) is
independent of computability: it holds for every codomain admitting a
fixed-point-free endomorphism, regardless of cardinality or
decidability.  The topological strand
(Theorem~\ref{thm:generic}) is likewise orthogonal to G\"odel-style
reasoning: it is a finite-data invariant, not an incompleteness of a
deductive system.

Section~\ref{sec:godel-distinction} provides a detailed side-by-side
comparison between G\"odel incompleteness and representational
incompleteness, addressing six independent structural differences and
explaining why the ``G\"odel repackaging'' objection fails on every
count.  The comparison also highlights the computability-independence
of the present result---it holds for all functions, computable or
not---which places it in a strictly larger domain than G\"odel- or
Turing-style arguments, which require recursive axiomatizability or
computability as a precondition.

\subsection{AI self-models and AGI limits}
Theoretical limits on AI self-understanding have been discussed from
computability-theoretic (Rice's theorem, Kleene's recursion theorem)
and information-theoretic perspectives.  The present work differs:
the diagonal exclusion (Theorem~\ref{thm:summit}) requires \emph{no
computability assumption}---it applies to every
$s : A \!\to\! A \!\to\! B$ regardless of computability---and the
topological barrier is orthogonal to recursion-theoretic machinery.
Full implications are developed in \S\ref{sec:ai-agi}.

\subsection{Topology in philosophy of mind}
Topological figures of speech appear in some philosophy of mind.  The present
work differs in two ways: (i)~its central topological claims are machine-checked
theorems rather than metaphors, and (ii)~the generalization is explicit---a
general lemma defining an open-ended class of obstructed space-pairs, not a
single hard-wired example.

% ============================================================================
% 10. CONCLUSION
% ============================================================================
\section{Conclusion}
\label{sec:conclusion}

The summit theorem (\leanref{representational\_incompleteness},
Theorem~\ref{thm:summit}) states the central result:
\textbf{no self-modeling system can capture its own diagonal}.  For any
$s : A \to A \to B$ and any fixed-point-free $f : B \to B$, the
diagonal function $a \mapsto f(s\,a\,a)$ is never equal to any row
$s\,a_0$---and if $s$ is claimed to be total and surjective, it simply
cannot exist.  No assumption on the surjectivity, cardinality, or
computability of $s$ is required.  This is
\textbf{representational incompleteness}: the analogue of G\"odel
incompleteness for self-modeling systems rather than formal theories.

Three mutually reinforcing layers make the result robust:

\begin{enumerate}[label=(\roman*)]
\item \textbf{Categorical infrastructure.}\enspace
  A machine-checked regress and Lawvere scaffold, including the
  \leanref{OntologicalSlot} wall that persists at every iterate
  depth, ensures that no climb up the meta-level tower dissolves the
  endo/obj distinction.
\item \textbf{Topological obstruction (general lemma + minimal witness).}\enspace
  Theorem~\ref{thm:generic} defines an \emph{open-ended class} of
  obstructed space-pairs: any two spaces with incompatible
  $\mathrm{ChartableR2}$-co-characterized boundaries are not
  homeomorphic.  The cylinder/M\"obius pair is the fully discharged
  \emph{minimal-case witness}; Klein bottles, projective planes, or
  any other surface satisfying the hypotheses would serve equally.
  The obstacle is \textbf{wrong type}, not
  \textbf{inexhaustible queue}.
\item \textbf{Systematic no-escape closure (Horns~1--6).}\enspace
  Six adversarial retreat routes are formally closed
  (\S\ref{sec:noescape}): infinite regress, finite-order collapse,
  cross-object readout, partial self-models, model substitution, and
  decidable self-reference.  Each closure is a proved theorem in
  Lean~4; together they leave an adversary with no viable route to
  circumvent either the diagonal or the topological barrier.
\end{enumerate}

\begin{keybox}[title=Three pillars of representational incompleteness]
\begin{enumerate}[label=\textbf{Pillar~\arabic*.},leftmargin=3.5em,itemsep=0.6ex]
\item \textbf{Diagonal exclusion}
  (Theorem~\ref{thm:summit}).  Every self-model $s : A \to A \to B$
  has a diagonal blind spot: the function $a \mapsto f(s\,a\,a)$ is
  never any row $s\,a_0$.  No assumption on surjectivity,
  computability, or cardinality.
\item \textbf{Topological obstruction}
  (Theorem~\ref{thm:generic}).  Any two spaces with incompatible
  $\mathrm{ChartableR2}$-detected boundaries are not homeomorphic.
  A finite-data invariant impervious to computation, scaling, or
  continuous deformation.
\item \textbf{Ontological sort wall}
  (Theorem~\ref{thm:wall}).  The \texttt{OntologicalSlot}
  construction enforces that the self-representation morphism
  ($\mathrm{endo}$) is definitionally disjoint from any global
  element ($\mathrm{obj}$) at every iterate depth.  The Lawvere
  fixed point does not dissolve this wall.
\end{enumerate}
Together, these three pillars---diagonal, topological, and
ontological---leave the adversary with no viable route: scaling,
iterating, partial modeling, cross-object readout, model substitution,
and decidable restriction are all formally inadmissible.  The
remaining admissible paths all require \emph{structural change} to the
topological type of the substrate.  The closure properties
(\leanref{ClosureProperties}) prove the barrier is robust under
composition, reindexing, post-composition, embedding, product
formation, multi-agent delegation, and nested simulation.
\end{keybox}

\paragraph{What is not being claimed.}
The theorems are unconditional mathematics; the \emph{application} to
any particular domain requires domain-specific premises (P2 for
consciousness, P3 for computational self-models,
\texttt{ChartableR2} biconditionals for a specific
state-space model, etc.).  The formalism does not prove that
consciousness is non-representational, that AI cannot be conscious, or
that symbol grounding is impossible.  It proves that \emph{if} the
relevant state space has the boundary pattern characterized here,
\emph{then} no homeomorphism to a space with a different boundary
pattern exists---and that the six most natural attempts to route around
that barrier fail in formally checkable ways.

\paragraph{The admissibility picture.}
The horn closures (\S\ref{sec:noescape}) together with the
admissibility analysis (\S\ref{sec:admissibility}) leave a clean
landscape.  All three pillars---diagonal (Pillar~1), topological
(Pillar~2), and ontological sort (Pillar~3)---contribute to the
closure.  In each neighboring debate---consciousness,
symbol grounding, diagonal limitations, measurement chains, AI
scaling, AGI self-improvement, the simulation hypothesis, and
computational theories of physics and mind---the admissible solutions
share a common shape:
\emph{structural change} to the topological type of the substrate,
not \emph{quantitative extension} within a fixed type.  Scaling,
iterating, partial modeling, finite-depth loops, cross-object readout,
and decidable restriction are all formally inadmissible.

\paragraph{What fills the blind spot.}
The blind spot is not a mere absence.  A functioning self-modeling
system \emph{does} produce diagonal behavior: it processes
self-referential inputs and generates outputs.  But no row of its
self-model matches that behavior
(\S\ref{sec:blind-spot-anatomy}).  Therefore, the system's diagonal
behavior is generated by substrate-level processes that the
self-model cannot represent.  The substrate is the \textbf{formally
necessary complement} to any parametric self-model.  Self-awareness
is not impossible---it occurs---but it occurs \emph{with} a diagonal
opacity, not without one.  The blind spot is the formal signature of
the substrate's irreducible contribution.  Perfectly self-transparent
self-awareness is the one thing that is structurally impossible.

\begin{keybox}[title={The final picture}]
Every self-modeling system is provably incomplete with respect to its
own diagonal.  Every attempt to close the gap---scaling, iterating,
delegating, restricting, embedding, changing the target
topology---is formally blocked.  Yet the system functions.
Therefore, \textbf{something outside the self-model generates the
diagonal behavior}: the substrate.  This is not optional; it is a
deductive consequence of the theorem and the empirical fact that
self-modeling systems exist and operate.  Representational
incompleteness does not say self-awareness is impossible.  It says
self-awareness is \textbf{irreducibly substrate-dependent}: the
self-model is a map, but the territory at the diagonal is charted
only by the substrate itself.  The gap between knower and known is
not a deficiency to be engineered away.  It is a proved,
permanent, structural feature of any system that turns its gaze
upon itself.
\end{keybox}

\noindent
The formalism makes this precise and machine-checkable, so that future
claims about circumventing the barrier can be evaluated against a
fixed standard rather than debated informally.

\medskip
\noindent\textbf{Future work.}
\begin{itemize}[leftmargin=1.5em,itemsep=0.15ex,topsep=0.3ex,parsep=0pt]
\item Discharge additional witnesses (Klein bottle, projective plane) for
  the general obstruction class.
\item A full $\mathrm{IsManifold}$ instance for
  $\mathrm{MobiusStrip}$.
\item A more abstract Lawvere statement for general cartesian closed
  categories.
\item Native higher-categorical horn closures ($2$-category and
  $\infty$-category analogues).
\item A full Turing-degree analogue of the decidable Lawvere horn
  (Rice's theorem / Kleene recursion in the same framework).
\end{itemize}

% ============================================================================
% APPENDICES
% ============================================================================
\appendix

\section{Lean Module Map}
\label{app:modules}

\footnotesize
\setlength{\tabcolsep}{3pt}
\begin{longtable}{@{}>{\raggedright\arraybackslash}p{3.95cm}>{\raggedright\arraybackslash}p{11.45cm}@{}}
\toprule
\textbf{Module} & \textbf{Role} \\
\midrule
\endfirsthead
\toprule
\textbf{Module} & \textbf{Role} \\
\midrule
\endhead
\midrule
\multicolumn{2}{r}{\textit{continued on next page}} \\
\endfoot
\bottomrule
\endlastfoot
\leanref{Basic} & \leanref{RepresentationalSystem}, iterates, \leanref{OntologicalSlot} \\
\leanref{Regress} & Non-termination, unbounded levels \\
\leanref{FixedPoints} & \leanref{EndoVsPoint}, Lawvere wall non-dissolution \\
\leanref{Lawvere} & Fixed-point theorem in \texttt{Type} \\
\leanref{LawvereCCCType} & \texttt{MonoidalClosed} packaging \\
\leanref{Orientability} & $T_2$ homeomorphism invariance \\
\leanref{CylinderMobius} & Fundamental domain, quotient, compact Hausdorff, boundary band \\
\leanref{CylinderBoundary} & Closed cylinder boundary: two disjoint connected faces \\
\leanref{CylinderChartableBoundary} & Cylinder C4 biconditional \\
\leanref{ChartableR2Neighbor} & \leanref{ChartableR2}, \leanref{ChartableR2BoundaryModel}; openness / stability \\
\leanref{ChartableR2Bridge} & \leanref{Homeomorph.chartableR2\_iff}; \leanref{not\_homeomorphic\_of\_chartableR2\_boundary\_contrast}; \leanref{ChartableR2BoundaryModel.not\_homeomorphic} \\
\leanref{ChartableR2ConcreteBoundaryModels} & M\"obius/cylinder boundary models; conditional observer chain; \leanref{QuantumObserverChainHypothesis} \\
\leanref{SymbolGrounding} & \leanref{SymbolSystem}; \leanref{symbolGrounding\_*} \\
\leanref{LawvereRegressBridge} & \leanref{no\_universal\_parametric\_unary\_nat}, \leanref{no\_universal\_parametric\_unary\_bool}; \leanref{lawvere\_diagonal\_not\_in\_range}; \leanref{lawvere\_fixed\_point\_stays\_representational} \\
\leanref{NoEscapeClosure} & Iterate collisions vs.\ injective towers; section--retraction idempotent readout \\
\leanref{ChartableR2Models} & Interior charts; cylinder interior \leanref{ChartableR2} \\
\leanref{MobiusChartableBoundary} & Open-cell charts; boundary $\neg$\leanref{ChartableR2}; M\"obius C4 packaging \\
\leanref{MobiusSeamChartable} & Saturated seam patches (open, saturated) \\
\leanref{MobiusSeamChart} & IFT-based local model; trig coordinates \\
\leanref{MobiusSeamTrigInject} & Trig injectivity on quotient windows \\
\leanref{MobiusSeamChartableR2} & Seam charts (trig + unfold); interior-height; \textbf{M-FINAL} \\
\leanref{MobiusCylinderBoundaryContrast} & Boundary subspaces not homeomorphic \\
\leanref{HalfLineVsLine} & 1D half-space $\neq$ line \\
\leanref{HalfPlaneVsPlane} & 2D half-space $\neq$ plane \\
\leanref{HalfSpaceNeighborVsPlane} & Local half-space $\neq$ open plane patch \\
\leanref{PuncturedPlaneNotSimplyConnected} & $\mathbb{C} \setminus \{0\}$ not simply connected \\
\leanref{OpenCompactChartObstruction} & Compact $\mathbb{R}^2$ patch $\neq$ $\mathbb{R}^2$ \\
\leanref{ClosureProperties} & Compositional inheritance; \leanref{ComputationalSelfModel} (P3); slot robustness; \leanref{TopologicalSelfModelingSystem} pillar bridge; exhaustive closure; simulation; computability independence \\
\leanref{Main} & Master theorem + extended bundle \\
\end{longtable}
\normalsize

\section{Key Theorem Index}
\label{app:theorems}

\footnotesize
\setlength{\tabcolsep}{3pt}
\begin{longtable}{@{}>{\raggedright\arraybackslash}p{3.85cm}>{\raggedright\arraybackslash}p{7.55cm}>{\raggedright\arraybackslash}p{3.15cm}@{}}
\toprule
\textbf{Lean name} & \textbf{Statement} & \textbf{Module} \\
\midrule
\endfirsthead
\toprule
\textbf{Lean name} & \textbf{Statement} & \textbf{Module} \\
\midrule
\endhead
\midrule
\multicolumn{3}{r}{\textit{continued on next page}} \\
\endfoot
\bottomrule
\endlastfoot
\leanref{representational\_incompleteness} & \textbf{Summit:} $\forall a_0$, $s\,a_0 \neq \lambda a.\,f(s\,a\,a)$; $\neg\exists$ total $s$ & \leanref{LawvereRegressBridge} \\
\midrule
\leanref{regress\_iterates\_unbounded} & $\forall B, \exists L > B$, distinct iterate & \leanref{Regress} \\
\leanref{lawvere\_fixed\_point\_Type} & Universal enumerator $\Rightarrow$ every endo has fixed point & \leanref{Lawvere} \\
\leanref{lawvere\_wall\_is\_not\_dissolution} & $\mathrm{endo}(\mathrm{represent}) \neq \mathrm{obj}(\mathrm{fp})$ & \leanref{FixedPoints} \\
\leanref{instT2SpaceMobiusStrip} & M\"obius strip is $T_2$ & \leanref{CylinderMobius} \\
\leanref{isConnected\_mobiusStripBoundary} & Boundary band is connected & \leanref{CylinderMobius} \\
\leanref{closedCylinder\_boundary\_eq\_union} & Cylinder boundary = 2 disjoint faces & \leanref{CylinderBoundary} \\
\leanref{euclideanHalfSpace2\_not\_homeomorphic\_euclidean2} & $H^2 \not\cong_t \mathbb{R}^2$ & \leanref{HalfPlaneVsPlane} \\
\leanref{not\_homeomorphic\_of\_chartableR2\_boundary\_contrast} & Generic: incompatible boundaries $\Rightarrow$ $\neg$ homeo & \leanref{ChartableR2Bridge} \\
\leanref{closedCylinder\_boundaryUnion\_iff\_not\_chartableR2} & Cylinder C4 biconditional & \leanref{CylinderChartableBoundary} \\
\leanref{not\_chartableR2\_of\_mem\_mobiusStripBoundary} & M\"obius boundary $\Rightarrow$ $\neg$\leanref{ChartableR2} & \leanref{MobiusChartableBdy} \\
\leanref{chartableR2\_mobiusQuotientMk\_of\_interior\_height} & Interior height $\Rightarrow$ \leanref{ChartableR2} & \leanref{MobiusSeamChartableR2} \\
\leanref{mobiusStripBoundary\_not\_homeomorphic\ldots} & Boundary subtypes not homeo & \leanref{MobiusCylBdyContrast} \\
\leanref{mobiusStrip\_not\_homeomorphic\_closedCylinder} & \textbf{M-FINAL} & \leanref{MobiusSeamChartableR2} \\
\leanref{representational\_regress\_master} & Core master conjunction & \leanref{Main} \\
\leanref{representational\_regress\_master\_extended} & Extended bundle (+ topology) & \leanref{Main} \\
\leanref{QuantumObserverChainHypothesis.not\_homeomorphic\_mobiusStrip} & Conditional: obs.\ chain $\not\simeq_t$ M\"obius & \leanref{ChartableR2ConcreteBoundaryModels} \\
\leanref{no\_universal\_parametric\_unary\_nat} & No universal $s : A \!\to\! A \!\to\! \mathbb{N}$ for all $g : A \!\to\! \mathbb{N}$ & \leanref{LawvereRegressBridge} \\
\leanref{IterateClosure.End.pow\_iterate\_dichotomy} & Injective iterate powers vs.\ finite index bound & \leanref{NoEscapeClosure} \\
\leanref{CrossObjectRepresentation.section\_retraction\_no\_injective\_tower\_either\_side} & Both-side: $\neg$ injective $\mathbb{N}$-powers on source \emph{or} host & \leanref{NoEscapeClosure} \\
\leanref{lawvere\_diagonal\_not\_in\_range} & Diagonal never in range of any $s : A \!\to\! A \!\to\! B$ (no surjectivity) & \leanref{LawvereRegressBridge} \\
\leanref{no\_universal\_parametric\_unary\_bool} & No universal $s : A \!\to\! A \!\to\! \mathrm{Bool}$ for all $g : A \!\to\! \mathrm{Bool}$ & \leanref{LawvereRegressBridge} \\
\midrule
\multicolumn{3}{l}{\textit{Closure properties (\leanref{ClosureProperties})}} \\
\leanref{ComputationalSelfModel.diagonal\_blind\_spot} & P3 bridge: self-model has diagonal blind spot & \leanref{ClosureProperties} \\
\leanref{inherited\_diagonal\_exclusion} & Subsystem blind spot propagates to containing system & \leanref{ClosureProperties} \\
\leanref{blind\_spot\_survives\_reindexing} & Reindexing via $r : C \to A$ preserves blind spot & \leanref{ClosureProperties} \\
\leanref{recursive\_self\_improvement\_blind\_spot} & Every link in an improvement chain has blind spot & \leanref{ClosureProperties} \\
\leanref{multiagent\_delegation\_blind\_spot} & Delegated model via $\pi : A_1 \!\to\! A_2$ has blind spot & \leanref{ClosureProperties} \\
\leanref{OntologicalSlot.wall\_survives\_injection} & Wall persists under injective relabeling & \leanref{ClosureProperties} \\
\leanref{blind\_spot\_survives\_postcomposition} & Post-composed model $h \circ s$ has blind spot & \leanref{ClosureProperties} \\
\leanref{crossObject\_both\_sides\_blind\_spot} & Embedding: A-side and C-side both have blind spot & \leanref{ClosureProperties} \\
\leanref{product\_self\_model\_blind\_spot} & Product of self-modeling systems has blind spot & \leanref{ClosureProperties} \\
\leanref{TopologicalSelfModelingSystem.both\_pillars\_apply} & Pillar bridge: diagonal + topological simultaneously & \leanref{ClosureProperties} \\
\leanref{combined\_diagonal\_and\_topological\_barrier} & Conjunction: blind spot $\wedge$ $\neg$surjective $\wedge$ $\neg$homeo & \leanref{ClosureProperties} \\
\leanref{pillar1\_full\_closure} & Full Pillar 1: diagonal + no surjection + $\mathbb{N}$ + Bool & \leanref{ClosureProperties} \\
\leanref{diagonal\_exclusion\_is\_constructive} & Diagonal exclusion needs no LEM & \leanref{ClosureProperties} \\
\end{longtable}
\normalsize

\section{Theorem Dependency Graph}
\label{app:deps}

The following diagram shows the logical dependency structure among the
principal theorems.  An arrow $A \to B$ means ``$B$ uses $A$ as a
lemma or hypothesis.''  The three pillars are shaded.

{\small
\noindent\textbf{Layer 1: Diagonal \& Categorical.}
\begin{itemize}[leftmargin=1.5em,itemsep=0.15ex,topsep=0.4ex,parsep=0pt]
\item \leanref{lawvere\_fixed\_point\_Type}
  $\to$
  diagonal-not-in-range, no-universal-nat, no-universal-bool\\
  $\to$ \textbf{[P1]}
  \leanref{representational\_incompleteness}
\item \leanref{OntologicalSlot} (def) $\to$
  \textbf{[P3]}
  \leanref{lawvere\_wall\_is\_not\_dissolution}
  $\to$
  \leanref{lawvere\_fixed\_point\_stays\_representational}
\item \leanref{regress\_iterates\_unbounded}
  $+$ P3 $\to$
  \leanref{representational\_regress\_master}
\item \leanref{pow\_iterate\_dichotomy} (Horn~2);\enspace
  section--retraction closure (Horn~3)
\end{itemize}

\noindent\textbf{Layer 2: Topological Obstruction.}
\begin{itemize}[leftmargin=1.5em,itemsep=0.15ex,topsep=0.4ex,parsep=0pt]
\item \leanref{Homeomorph.chartableR2\_iff} $\to$
  \textbf{[P2]}
  \leanref{not\_homeomorphic\_of\_chartableR2\_boundary\_contrast}
\item Cylinder C4 $+$ M\"obius C4 $+$ boundary contrast $\to$
  \leanref{mobiusStrip\_not\_homeomorphic\_closedCylinder}
  \textbf{(M-FINAL)}
\item Interior-height chartability $\to$ M\"obius C4;\enspace
  half-space obstruction $\to$ Cylinder C4
\end{itemize}

\noindent\textbf{Layer 3: Integration.}
\begin{itemize}[leftmargin=1.5em,itemsep=0.15ex,topsep=0.4ex,parsep=0pt]
\item \leanref{representational\_regress\_master}
  $+$ M-FINAL $\to$
  \leanref{representational\_regress\_master\_extended}
\end{itemize}

\noindent\textbf{Layer 4: Closure Properties.}
\begin{itemize}[leftmargin=1.5em,itemsep=0.15ex,topsep=0.4ex,parsep=0pt]
\item Summit $\to$
  inherited-diagonal, reindexing,
  post-composition, product-model,
  exhaustive-closure
\item \leanref{OntologicalSlot} $\to$
  wall-survives-injection,
  iterate-tower-injective
\item \leanref{ComputationalSelfModel} (P3)
  $\to$ diagonal-blind-spot,
  no-complete-self-model
\item cross-object $\to$
  both-sides-blind-spot,
  delegation-blind-spot
\item \leanref{TopologicalSelfModelingSystem}
  $\to$ both-pillars-apply,
  combined-barrier
\item pillar1/2/3-full-closure $+$
  categorical-horn-closures
\end{itemize}
}% end \small

\noindent
\textbf{Summary statistics.}
The Lean development contains 361 theorems and 98 lemmas across 30
source files, with zero \texttt{sorry} and no axioms beyond Lean~4's
standard foundation (propositional extensionality, quotients, choice).
The Key Theorem Index (Appendix~\ref{app:theorems}) lists the 35
principal results; all are verified present in the source.

\section{Reproduction}
\label{app:repro}

\begin{verbatim}
cd representational-incompleteness-lean
lake update
lake exe cache get
lake build RepresentationalIncompleteness
# Verify: 0 sorry
rg "sorry" RepresentationalIncompleteness/
\end{verbatim}

\noindent
Toolchain: \texttt{leanprover/lean4:v4.29.0-rc6}. Mathlib: \texttt{v4.29.0-rc6}.

\end{document}
