% The Representational Regress — LaTeX source (Lean 4 / Mathlib companion library)

\documentclass[11pt]{article}
\usepackage[margin=1in]{geometry}
\usepackage{amsmath,amssymb,amsthm}
\usepackage{hyperref}
\usepackage{enumitem}
\usepackage{array}
\usepackage{booktabs}
\usepackage{xcolor}
\usepackage{stmaryrd}

% Allow line breaks in long Lean identifiers (\_, namespaces, CamelCase).
\ExplSyntaxOn
\tl_new:N \l__rr_leanref_tl
\NewDocumentCommand \leanref { m }
  {
    \tl_set:Nn \l__rr_leanref_tl { #1 }
    \regex_replace_all:nnN { \c{_} } { \c{_} \c{allowbreak} } \l__rr_leanref_tl
    \regex_replace_all:nnN { \. } { . \c{allowbreak} } \l__rr_leanref_tl
    % CamelCase / digit-boundary splits (Lean module and declaration names).
    \regex_replace_all:nnN { ([a-z])([A-Z]) } { \1 \c{allowbreak} \2 } \l__rr_leanref_tl
    \regex_replace_all:nnN { ([0-9])([A-Z]) } { \1 \c{allowbreak} \2 } \l__rr_leanref_tl
    \texttt{\small \tl_use:N \l__rr_leanref_tl }
  }
\ExplSyntaxOff

\newtheorem{theorem}{Theorem}[section]
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{proposition}[theorem]{Proposition}
\newtheorem{corollary}[theorem]{Corollary}
\newtheorem{definition}[theorem]{Definition}
\newtheorem{remark}[theorem]{Remark}

\newcommand{\novel}{\textcolor{blue!70!black}{[Novel]}}
\newcommand{\known}{\textcolor{gray}{[Known math]}}
% Break `theorem`/`definition` head away from long \leanref tags (avoids overfull italic headers).
\newcommand{\TheoremLeanTag}[1]{%
  \noindent\textnormal{(\leanref{#1})}\par\noindent}
% End `amsthm` head paragraph so the declaration tag is not merged into the header line.
\newcommand{\AfterTheoremHead}{\mbox{}\par}

\title{The Representational Regress:\\
  Machine-Checked Topology, Diagonal Arguments,\\
  and Boundary Obstructions for Representational Systems}
\author{Nova Spivack\\
Independent Researcher}
\date{2026}

\begin{document}
\maketitle

% ============================================================================
% ABSTRACT
% ============================================================================
\begin{abstract}
This work is a Lean~4 formalization, checked against Mathlib, with zero
\texttt{sorry} and no project-specific axioms beyond the Lean core. It treats
\emph{representational systems}---categories equipped with a non-trivial iterated
self-map---and an intrinsic planar-chart predicate \texttt{ChartableR2} on compact
topological models.

\begin{enumerate}[label=(\roman*)]
\item \textbf{Categorical strand.} A representational system yields infinitely
  many formally distinct meta-levels; Lawvere's fixed-point theorem in
  \texttt{Type} and in \texttt{MonoidalClosed(Type)} is complemented by an
  inductive \texttt{OntologicalSlot} witness that separates morphism data from
  global-element data, so the ``Lawvere wall'' does not collapse that
  distinction inside the formalism.
\item \textbf{Topological strand.} The closed cylinder $S^1 \times [0,1]$ and the
  M\"obius strip (quotient of $[0,1]^2$ by $(0,t)\sim(1,1-t)$) are not
  homeomorphic. Points that admit an open neighborhood homeomorphic to
  $\mathbb{R}^2$ are classified by \texttt{ChartableR2}; on both models the
  chosen boundary agrees with ``not \texttt{ChartableR2},'' while the boundary
  \emph{subspaces} differ (connected vs.\ disconnected). A general lemma,
  \texttt{not\_homeomorphic\_of\_chartableR2\_boundary\_contrast}, states the
  same obstruction for any two spaces once those biconditionals and a subtype
  non-homeomorphism are supplied.
\item \textbf{Bridges to neighboring debates.} The formal library also
  includes \texttt{SymbolSystem} (synonymous with \texttt{RepresentationalSystem})
  and lemmas restating regress injectivity in the vocabulary of interpretation;
  \texttt{LawvereRegressBridge}, which derives impossible ``universal
  parametric'' enumerations of unary maps and relates Lawvere phenomena to
  \texttt{OntologicalSlot}; bundled \texttt{ChartableR2BoundaryModel} instances
  for the cylinder and M\"obius strip; and \texttt{QuantumObserverChainHypothesis},
  a structure whose fields are exactly the premises of the general boundary
  lemma---intended for a \emph{hypothetical} topological model of an
  observer--readout chain, without encoding Hilbert spaces or measurement axioms
  in the present development.
  No theorem refers to consciousness; a supplementary philosophical premise~P2,
  that a target notion of self-presence is modeled globally by M\"obius-type
  rather than cylinder-type organization, combines with~(ii) to yield a
  conditional reading under which representational substrates matching the
  cylinder-side boundary pattern cannot be homeomorphic to a space matching the
  M\"obius-side pattern---a gap that, when those premises apply, is not removed by
  scaling, iteration, or continuous deformation within a fixed topological type.
\item \textbf{No-escape packaging}\newline
  \noindent(\leanref{NoEscapeClosure},
  \leanref{LawvereRegressBridge}).\newline
  \noindent A separate strand makes adversarial ``horn'' moves explicit,
  closing six routes by which an opponent might claim to circumvent
  the regress or the Lawvere wall:
  \begin{itemize}[leftmargin=1.5em,itemsep=0pt,topsep=0.3ex,parsep=0pt]
  \item[\emph{(a)}] Iterate collisions force a \emph{finite} repertoire
    of monoid powers---no infinite injective $\mathbb{N}$-tower along a
    periodic channel.
  \item[\emph{(b)}] A section--retraction readout induces an idempotent
    loop on the host and the identity on the source, blocking injective
    towers on \emph{both} sides.
  \item[\emph{(c)}] Even \emph{partial} (non-surjective) self-models
    necessarily miss their own diagonal---no surjectivity
    assumption required.
  \item[\emph{(d)}] A decidable/computability analogue closes the finite
    Lawvere horn using $\lnot : \mathrm{Bool}\to\mathrm{Bool}$.
  \end{itemize}
  These lemmas do not replace \texttt{iter\_injective}, but spell out
  what happens when a rival weakens, reroutes, or partially implements
  self-representation.
\end{enumerate}

\medskip
\noindent The source tree contains roughly twenty-five Lean modules; the body
and appendices summarize the \texttt{OntologicalSlot} construction, seam and
equator charts for \texttt{ChartableR2}, scope, and reproduction.
\end{abstract}

% ============================================================================
% 1. INTRODUCTION
% ============================================================================
\section{Introduction}
\label{sec:intro}

\subsection{The Problem}
Representational systems---symbolic AI, formal theories, or any setting where a
structure represents itself by iterating a fixed construction---support arbitrarily
deep \emph{meta}-levels. Classical diagonal and fixed-point arguments constrain
how much ``self-capture'' such a layer can achieve. Independently, one may ask
whether intrinsic topology distinguishes different \emph{kinds} of ``inside vs.\
outside'' organization when global homeomorphism is taken as geometric
sameness.

This paper gives a Lean~4 formalization of a sizeable fragment of that joint
landscape. The technical development stays in categorical and topological
language; \S\ref{sec:instantiations} relates the results to consciousness,
symbolic grounding, diagonal arguments, and a template for observer--apparatus
topologies.

\subsection{Summary of Results}
The formalization proves, with zero \texttt{sorry} and no custom axioms, a
\textbf{layered} constraint on representational systems. The layers are
logically independent: one can accept the topological results without leaning on
an infinite tower, and one can study the tower without the cylinder/M\"obius
contrast. Together they imply that ``more representation'' is not a generic path
from one global geometric type to another.

\begin{enumerate}[label=(\alph*)]
\item \textbf{Categorical regress and Lawvere wall} (\S\ref{sec:regress}):
  under a bundled non-collapse hypothesis on iterates of \texttt{represent},
  there are infinitely many formally distinct meta-levels; Lawvere's theorem in
  \texttt{Type} / CCC form is complemented by \texttt{OntologicalSlot}, so a
  fixed-point tally cannot erase the morphism-vs-global-element typing.
\item \textbf{Universal topological obstruction} (\S\ref{sec:generic},
  \S\ref{sec:mfinal}): Theorem~\ref{thm:generic} states a \emph{template}---if two
  spaces admit subsets characterized as exactly the non-\texttt{ChartableR2}
  points, and if those boundary \emph{subspaces} are not homeomorphic, then the
  ambient spaces are not homeomorphic. \textbf{M-FINAL} is a fully worked
  compact witness (cylinder vs.\ M\"obius strip). Any future domain whose
  state space can be modeled with the same sensor pattern and an incompatible
  boundary subtype imports the same conclusion; no separate bridge argument is
  required.
\item \textbf{Extended master bundle} (\S\ref{sec:extended}):
  \texttt{representational\_regress\_master\_extended} adjoins the $1$D
  half-space non-homeomorphism and M-FINAL to the core master conjunction.
\item \textbf{Cross-domain bridges} (\S\ref{sec:instantiations}):
  \begin{itemize}[leftmargin=1.75em,itemsep=0.15ex,topsep=0.5ex,parsep=0pt]
  \item \texttt{SymbolSystem}
  \item \texttt{LawvereRegressBridge}
  \item concrete \texttt{ChartableR2BoundaryModel} instances
  \item \texttt{QuantumObserverChainHypothesis}
  \end{itemize}
  \noindent These make the obstruction's \emph{hypothesis shape} explicit for
  grounding, diagonal, and observer--readout readings.
\item \textbf{Adversarial horn closures}
  (\S\ref{sec:regress}, \S\ref{sec:noescape}).\newline
  \noindent Six escape routes are formally blocked in
  \leanref{NoEscapeClosure} and
  \leanref{LawvereRegressBridge}: iterate collision dichotomy,
  both-sided section--retraction idempotent collapse,
  diagonal exclusion for partial self-models (no surjectivity
  required), decidable Lawvere for $\mathrm{Bool}$-valued functions,
  and the original $\mathbb{N}$-valued enumerator obstruction.
  Together these leave an adversary with a clean dichotomy: either
  an infinite injective regress with Lawvere wall intact, or a finite
  loop too shallow for full self-capture.
\end{enumerate}

\paragraph{Significance and what is \emph{not} being claimed.}
Much informal discussion stops at \emph{infinite regress}: every
meta-representation invites another, so a desired ``closure'' fails. That
phenomenon is real in the formalism (item~(a)), but it is not the deepest
punchline. The topological strand (item~(b)) is \textbf{not} a complaint about
non-termination: homeomorphism obstructions are \emph{finite-data invariants} in
spirit---two spaces are either homeomorphic or not, regardless of whether one
enumerates infinitely many modeling steps. The barrier is that \emph{no}
continuous global identification (a homeomorphism) can exist between disparate
types once the \texttt{ChartableR2}-boundary pattern and subtype mismatch are in
place. In particular, the obstacle is not removed by ``more compute,'' longer
iterates, richer self-models, or architectural reparameterization if those
changes do not alter the underlying topological type of the modeled substrate.
Infinity can distract from the point; the point is \textbf{wrong type}, not
\textbf{inexhaustible queue}.

\paragraph{Utility.}
For any application---symbolic AI, formal theories, measurement chains, or
philosophical models---the workflow is: (1)~encode the system (or its
state space) in sufficient topological detail to verify the boundary biconditionals,
(2)~compare boundary subspaces up to homeomorphism, (3)~invoke the general
lemma or a bundled model. If the premises hold, the obstruction applies; if they
fail, the obstacle was never licensed. That conditional structure is what keeps
the mathematics honest while still delivering a \emph{portable} gate.

\subsection{Contributions and Novelty Claims}
\begin{enumerate}[label=(\arabic*)]
\item \textbf{\texttt{OntologicalSlot}} --- inductive tagging of morphism
  slots versus global-element slots; wall lemmas follow from constructor
  disjointness yet make the Lawvere-adjacent moral explicit inside the type
  discipline. [\novel]
\item \textbf{\texttt{ChartableR2} spine + seam--equator charts} --- intrinsic
  $\mathbb{R}^2$ chartability as a definable ``planar interior vs.\ boundary''
  sensor; constructive charts through the quotient seam including the equator via
  piecewise unfolding. [\novel\ application / technique]
\item \textbf{General boundary-contrast obstruction.}\newline
  \texttt{not\_homeomorphic\_of\_chartableR2\_boundary\_contrast}\newline
  \noindent A single theorem that \emph{packages} the logical move used in
  M-FINAL: any two
  spaces hit the same wall once their \texttt{ChartableR2}-co-characterized
  boundaries are topologically incompatible. Individual proofs of
  non-homeomorphism are classical; a \emph{reusable, machine-checked} lemma with
  this generality, wired to a concrete compact witness and to a categorical
  regress bundle, is---to our knowledge---\emph{not} standard in the
  philosophy-of-mind or AI-safety literatures (where topology more often appears
  metaphorically). We claim distinctiveness for the \textbf{integrated,
  checkable} template, not for the elementary fact that the M\"obius strip and
  the cylinder differ. [\novel]
\item \textbf{Supplementary structures:}
  \begin{itemize}[leftmargin=1.75em,itemsep=0.15ex,topsep=0.5ex,parsep=0pt]
  \item \texttt{SymbolSystem}
  \item \texttt{LawvereRegressBridge}
  \item bundled \texttt{ChartableR2BoundaryModel} data
  \item \texttt{QuantumObserverChainHypothesis}
  \end{itemize}
  \noindent These spell out which hypotheses must be imported from a domain
  before the obstruction applies; they do not expand the set of unconditional
  theorems but sharpen \emph{where} the gate sits. [\novel]
\item \textbf{\leanref{NoEscapeClosure} + extended
  \leanref{LawvereRegressBridge}}\newline
  \noindent Packages six complementary ``escape routes,'' each formally
  blocked:
  \begin{itemize}[leftmargin=1.5em,itemsep=0pt,topsep=0.3ex,parsep=0pt]
  \item[(i)] Iterate collision $\Rightarrow$ finite power repertoire.
  \item[(ii)] Section--retraction readout $\Rightarrow$ idempotent on
    host, identity on source; no injective tower on either side.
  \item[(iii)] Partial (non-surjective) self-model $\Rightarrow$
    diagonal blind spot.
  \item[(iv)] Decidable evaluator into $\mathrm{Bool}$ $\Rightarrow$
    no universal parametric unary.
  \item[(v)] Full Lawvere for $\mathbb{N}$ and general $B$ with
    fixed-point-free endo.
  \item[(vi)] Wall persists at every iterate depth.
  \end{itemize}
  [\novel]
\item \textbf{Classical ingredients.} Lawvere's diagonal argument (1969),
  textbook topology of the cylinder and M\"obius strip, and Cantorian regress
  patterns precede this work. The contribution is a \textbf{unified,
  adversarially checked} landscape in which (i)~regress and Lawvere infrastructure,
  (ii)~a \emph{general} topological obstruction, and (iii)~a discharged witness
  and extension hooks coexist in one artifact, so that ``compute harder'' is
  not confusable with ``change homeomorphism type.''
\end{enumerate}

\subsection{Paper Organization}
Section~\ref{sec:regress}: representational systems, regress, Lawvere,
\texttt{OntologicalSlot}, core master theorem.
Section~\ref{sec:models}--\ref{sec:mfinal}: cylinder and M\"obius models,
\texttt{ChartableR2}, generic lemma, seam charts, M-FINAL.
Section~\ref{sec:extended}: extended master and instantiations (consciousness,
grounding, diagonal arguments, QM schematic).
Sections~\ref{sec:limits}--\ref{sec:conclusion}: limits, related work, conclusion.
Appendices: Lean module map, theorem index, reproduction.

% ============================================================================
% 2. CATEGORICAL STRAND: THE REPRESENTATIONAL REGRESS
% ============================================================================
\section{The Representational Regress}
\label{sec:regress}

\subsection{The \texttt{RepresentationalSystem} Structure}
\label{sec:repsys}

\begin{definition}[\leanref{Basic.lean}: \leanref{RepresentationalSystem}]
\leavevmode\par\noindent
A \emph{representational system} is a tuple $(C, A, \mathrm{represent}, \mathrm{iter\_injective})$ where:
\begin{itemize}
\item $C$ is a category,
\item $A$ is an object of $C$,
\item $\mathrm{represent} : A \to A$ is an endomorphism,
\item $\mathrm{iter\_injective}$: for all $n \neq m$, $\mathrm{represent}^n \neq \mathrm{represent}^m$ in $\mathrm{End}(A)$.
\end{itemize}
\end{definition}

\subsection{Non-Termination and Unboundedness}

\begin{theorem}[\known\ \leanref{Regress.lean}]
\label{thm:regress}
For any representational system $R$:
\begin{enumerate}[label=(\alph*)]
\item (\leanref{regress\_no\_termination}) Distinct indices $n \neq m$ yield distinct meta-level arrows.
\item (\leanref{regress\_iterates\_unbounded}) For every bound $B$, there exists a level $L > B$ with a distinct iterate.
\item (\leanref{regress\_is\_infinite}, \leanref{meta\_range\_infinite}) The set of meta-levels is infinite.
\end{enumerate}
\end{theorem}

\begin{remark}[What if \texttt{iter\_injective} fails?]
\label{rem:dichotomy}
The hypothesis $\mathrm{iter\_injective}$ is \emph{bundled}, not derivable from
a bare endomorphism. An adversary might weaken it: suppose
$\mathrm{represent}^n = \mathrm{represent}^m$ for some $n \neq m$.
\leanref{NoEscapeClosure} formalizes the consequence: let
$e := \mathrm{End.of}(\mathrm{represent})$; if $e^{n+d}=e^n$ with $d>0$ then
every power $e^k$ equals some $e^i$ with $i < n+d$
(\leanref{IterateClosure.End.exists\_lt\_of\_pow\_collision}).
\leanref{IterateClosure.End.pow\_iterate\_dichotomy}
records the \textbf{dichotomy}: for every monoid element~$e$, either
(i)~distinct natural-number powers are all distinct (an infinite injective
$\mathbb{N}$-tower), or (ii)~the set of \emph{distinct} powers is bounded by a
finite index.  In case~(ii) the system has only finitely many representational
layers---it loops.  Either route blocks ``just enough'' self-representation: one
gives an infinite regress; the other gives a finite loop too impoverished to
host injective iterate depth.
\end{remark}

\subsection{The Lawvere Wall}
\label{sec:lawvere}

\begin{theorem}[\known\ \leanref{Lawvere.lean}]
\label{thm:lawvere}
(\leanref{lawvere\_fixed\_point\_Type}) If $s : A \to (A \to B)$ is surjective, then every $f : B \to B$ has a fixed point.
\end{theorem}

\begin{theorem}[\known\ \leanref{LawvereCCCType.lean}]
\label{thm:lawvere-ccc}
(\leanref{lawvere\_fixed\_point\_MonoidalClosedType}) The same conclusion holds when the universal-enumerator hypothesis is recast as surjectivity of $\mathrm{curry}(\mathrm{lawvereBinary}\ s)$ in $\mathrm{MonoidalClosed}(\mathrm{Type}\ u)$.
\end{theorem}

\subsection{The \texttt{OntologicalSlot} Construction}
\label{sec:ontoslot}

\begin{definition}[\novel\ \leanref{Basic.lean}: \leanref{OntologicalSlot}]
\label{def:ontoslot}
For a category $C$ with terminal object and an object $A$:
\[
\mathrm{OntologicalSlot}(C, A) \;:=\;
  \mathrm{endo}(f : A \to A) \;\mid\; \mathrm{obj}(g : \top_C \to A)
\]
an inductive type with two constructors, disjoint by definition.
\end{definition}

\begin{theorem}[\novel\ \leanref{FixedPoints.lean}]
\label{thm:wall}
(\leanref{lawvere\_wall\_is\_not\_dissolution}) For any representational system $R$ and any global element $\mathrm{fp} : \top_C \to A$,
\[
  \mathrm{OntologicalSlot.endo}(R.\mathrm{represent}) \;\neq\; \mathrm{OntologicalSlot.obj}(\mathrm{fp}).
\]
\end{theorem}

\subsection{The Core Master Theorem}

\begin{theorem}[\leanref{Main.lean}]
\label{thm:master}
(\leanref{representational\_regress\_master}) For any representational system $R$:
\[
  \mathrm{representational\_regress\_master\_claim}(R)
\]
i.e., the conjunction of unbounded regress, Lawvere wall non-dissolution, and $T_2$ invariance under homeomorphism.
\end{theorem}

% ============================================================================
% 3. TOPOLOGICAL MODELS
% ============================================================================
\section{Topological Models: Cylinder and M\"obius Strip}
\label{sec:models}

\subsection{The Closed Cylinder}
\label{sec:cylinder}

\begin{definition}[\leanref{CylinderBoundary.lean}]
$\mathrm{ClosedCylinder} := S^1 \times [0,1]$ with the product topology and manifold structure $\mathfrak{R}_1 \times \mathfrak{R}_1^\partial$.
\end{definition}

\begin{theorem}[\known\ \leanref{CylinderBoundary.lean}]
\label{thm:cyl-boundary}
The manifold boundary of \emph{ClosedCylinder} decomposes as:
\[
  \partial(\mathrm{ClosedCylinder}) = F_{\mathrm{bot}} \sqcup F_{\mathrm{top}}
\]
where $F_{\mathrm{bot}} = S^1 \times \{0\}$ and $F_{\mathrm{top}} = S^1 \times \{1\}$ are each connected, and their union is \emph{disconnected}.
\end{theorem}

\subsection{The M\"obius Strip}
\label{sec:mobius}

\begin{definition}[\leanref{CylinderMobius.lean}]
\label{def:mobius-rel}
The \emph{M\"obius equivalence} $\sim$ on $[0,1]^2$ is the equivalence relation generated by
$(0, t) \sim (1, 1-t)$ for $t \in [0,1]$, extended by reflexivity, symmetry, and transitivity.
The \emph{M\"obius strip} is $\mathrm{MobiusStrip} := [0,1]^2 / {\sim}$.
\end{definition}

\begin{theorem}[\leanref{CylinderMobius.lean}]
\label{thm:mobius-props}
\leavevmode
\begin{enumerate}[label=(\alph*)]
\item (\leanref{mobiusRel0Graph\_isClosed}) The graph of $\sim$ is closed in $[0,1]^2 \times [0,1]^2$.
\item (\leanref{CompactSpace MobiusStrip}) $\mathrm{MobiusStrip}$ is compact (quotient of compact).
\item (\leanref{instT2SpaceMobiusStrip}) $\mathrm{MobiusStrip}$ is Hausdorff (closed graph + compact domain).
\item (\leanref{isConnected\_mobiusStripBoundary}) The boundary band $\partial_{\mathrm{model}} := \pi(\{t = 0\} \cup \{t = 1\})$ is connected (one component).
\end{enumerate}
\end{theorem}

\subsection{Boundary Subspace Contrast}

\begin{theorem}[\leanref{MobiusCylinderBoundaryContrast.lean}]
\label{thm:boundary-contrast}
\AfterTheoremHead
\TheoremLeanTag{mobiusStripBoundary\_not\_homeomorphic\_closedCylinderBoundaryUnion}
\[
  \mathrm{IsEmpty}\!\left(\mathrm{Subtype}(\partial_{\mathrm{Mob}}) \simeq_t \mathrm{Subtype}(\partial_{\mathrm{Cyl}})\right).
\]
The M\"obius boundary band (connected) and the cylinder boundary union (disconnected) are not homeomorphic as subspaces.
\end{theorem}

% ============================================================================
% 4. THE CHARTABLER2 FRAMEWORK
% ============================================================================
\section{The \texttt{ChartableR2} Framework}
\label{sec:chartable}

\subsection{Definition and Basic Properties}

\begin{definition}[\leanref{ChartableR2Neighbor.lean}]
\label{def:chartable}
A point $x$ of a topological space $X$ is \emph{$\mathrm{ChartableR2}$} if there exists an open set $U \ni x$ such that $U \cong_t \mathbb{R}^2$.
\end{definition}

\begin{theorem}[\leanref{ChartableR2Neighbor.lean}]
\label{thm:chartable-props}
\leavevmode
\begin{enumerate}[label=(\alph*)]
\item (\leanref{isOpen\_setOf\_chartableR2}) The set of \texttt{ChartableR2} points is open.
\item (\leanref{isClosed\_setOf\_not\_chartableR2}) Its complement is closed.
\item (\leanref{not\_chartableR2\_of\_tendsto\_seq\_atTop}) Sequential stability: limits of non-chartable points are non-chartable.
\end{enumerate}
\end{theorem}

\subsection{Invariance Under Homeomorphism}

\begin{theorem}[\leanref{ChartableR2Bridge.lean}]
\label{thm:chartable-invar}
\AfterTheoremHead
\TheoremLeanTag{Homeomorph.chartableR2\_iff}
If $h : X \simeq_t Y$ is a homeomorphism, then $\mathrm{ChartableR2}(x) \iff \mathrm{ChartableR2}(h(x))$.
\end{theorem}

\subsection{The Generic Boundary-Contrast Obstruction}
\label{sec:generic}

\begin{theorem}[\novel\ \leanref{ChartableR2Bridge.lean}]
\label{thm:generic}
\AfterTheoremHead
\TheoremLeanTag{not\_homeomorphic\_of\_chartableR2\_boundary\_contrast}
Let $X$ and $Y$ be topological spaces with predicates $b_X \subseteq X$ and $b_Y \subseteq Y$ such that:
\begin{enumerate}[label=(\roman*)]
\item $\forall x,\; x \in b_X \iff \neg\,\mathrm{ChartableR2}(x)$,
\item $\forall y,\; y \in b_Y \iff \neg\,\mathrm{ChartableR2}(y)$,
\item $\mathrm{IsEmpty}(\mathrm{Subtype}(b_X) \simeq_t \mathrm{Subtype}(b_Y))$.
\end{enumerate}
Then $\mathrm{IsEmpty}(X \simeq_t Y)$.
\end{theorem}

\begin{proof}[Proof sketch]
Any homeomorphism $e : X \simeq_t Y$ preserves $\mathrm{ChartableR2}$
(Theorem~\ref{thm:chartable-invar}), hence maps $b_X$ to $b_Y$ pointwise.
By $e.\mathrm{subtype}$, we get a homeomorphism
$\mathrm{Subtype}(b_X) \simeq_t \mathrm{Subtype}(b_Y)$, contradicting~(iii).
\end{proof}

Theorem~\ref{thm:generic} subsumes the non-homeomorphism strategy of \S\ref{sec:mfinal}:
any other pair of spaces satisfying the same three hypotheses is also not homeomorphic.

\subsection{Half-Space Obstructions}
\label{sec:halfspace}

\begin{theorem}[\leanref{HalfPlaneVsPlane.lean}]
\label{thm:halfplane}
\AfterTheoremHead
\TheoremLeanTag{euclideanHalfSpace2\_not\_homeomorphic\_euclidean2}
$\mathrm{IsEmpty}(H^2 \simeq_t \mathbb{R}^2)$.
Proof via punctured half-plane (star-convex $\Rightarrow$ simply connected) vs.\ punctured plane (not simply connected, from $\exp$ covering).
\end{theorem}

\begin{theorem}[\leanref{HalfSpaceNeighborVsPlane.lean}]
\label{thm:halfspace-local}
\AfterTheoremHead
\TheoremLeanTag{isEmpty\_homeomorph\_subtype\_openH2\_zero\_nbhd\_subtype\_openPlane}
For open $V \subseteq H^2$ with $0 \in V$ and open $O \subseteq \mathbb{R}^2$, there is no homeomorphism $V \cong_t O$.
\end{theorem}

\subsection{Cylinder C4: Boundary Biconditional}

\begin{theorem}[\leanref{CylinderChartableBoundary.lean}]
\label{thm:cylinder-c4}
\AfterTheoremHead
\TheoremLeanTag{closedCylinder\_boundaryUnion\_iff\_not\_chartableR2}
For all $x \in \mathrm{ClosedCylinder}$:
\[
  x \in \partial_{\mathrm{Cyl}} \;\iff\; \neg\,\mathrm{ChartableR2}(x).
\]
\end{theorem}

% ============================================================================
% 5. THE SEAM-CHART PROBLEM
% ============================================================================
\section{Interior Charts on the M\"obius Strip: The Seam Problem}
\label{sec:seam}

\subsection{Why the Seam Is Hard}
\label{sec:seam-why}

Interior-height points of the fundamental domain fall into three regimes: strict
interior $0<x<1$, where the quotient is injective on small boxes; the vertical
seam $x\in\{0,1\}$ with $t\neq\tfrac{1}{2}$, where trigonometric coordinates can
separate fibers when the height window avoids the equator; and the seam at
$t=\tfrac{1}{2}$, where $(0,\tfrac{1}{2})\sim(1,\tfrac{1}{2})$ and a trigonometric
chart degenerates. $\mathrm{ChartableR2}$ at the equator quotient class is
obtained by a piecewise Euclidean unfolding map (see below).

\subsection{Open-Cell Charts}

\begin{theorem}[\leanref{MobiusChartableBoundary.lean}]
\label{thm:open-cell}
\AfterTheoremHead
\TheoremLeanTag{chartableR2\_mobiusQuotientMk\_of\_Ioo\_square}
If $p \in (0,1)^2$ (strict interior of the fundamental domain), then $\mathrm{ChartableR2}(\llbracket p \rrbracket)$.
\end{theorem}

\subsection{Seam Trig Charts (Away from Equator)}

\begin{definition}[\leanref{MobiusSeamChart.lean}, \leanref{MobiusSeamChartable.lean}]
The \emph{saturated seam patch} $S(t_0, \varepsilon, \delta)$ is the union of the left and right $\varepsilon$-thickened seam strips at height window $(t_0 - \delta, t_0 + \delta)$, saturated under $\sim$.
\end{definition}

\begin{theorem}[\leanref{MobiusSeamTrigInject.lean}]
\label{thm:trig-inject}
\AfterTheoremHead
\TheoremLeanTag{onPatch\_mobiusFDTrigCoord\_eq\_iff\_mobiusRel0}
When $\delta < |t_0 - \tfrac{1}{2}|$, the trigonometric FD coordinate map $\mathrm{mobiusFDTrigCoord}$ satisfies:\newline
\noindent$\mathrm{mobiusFDTrigCoord}(p) = \mathrm{mobiusFDTrigCoord}(q)$ on the saturated patch if and only if $p \sim q$.
\end{theorem}

\begin{theorem}[\leanref{MobiusSeamChartableR2.lean}]
\label{thm:seam-trig}
\AfterTheoremHead
\TheoremLeanTag{chartableR2\_mobiusQuotientMk\_of\_mem\_mobiusSeamSaturatedPatch}
For $t_0 \neq \tfrac{1}{2}$ and appropriate $\varepsilon, \delta$, every point\newline
\noindent in $\pi(S(t_0, \varepsilon, \delta))$ is $\mathrm{ChartableR2}$.
\end{theorem}

\subsection{The Equator Chart: Piecewise Unfolding}
\label{sec:equator}

\begin{definition}[\leanref{MobiusSeamChartableR2.lean}]
\label{def:unfold}
\leavevmode\par\noindent
The \emph{fundamental-domain unfolding map}\newline
\noindent$\mathrm{unfoldFD} : [0,1]^2 \to \mathbb{R}^2$ is:
\[
  \mathrm{unfoldFD}(x, t) :=
  \begin{cases}
    \mathrm{mobiusPlaneCoord}(x, t) & \text{if } x \leq \tfrac{1}{2}, \\
    \begin{aligned}[t]
      &\mathrm{mobiusPlaneTrigSeamPartnerSub}\\
      &\quad\bigl(\mathrm{mobiusPlaneCoord}(x, t)\bigr)
    \end{aligned} & \text{if } x > \tfrac{1}{2}.
  \end{cases}
\]
\end{definition}

\begin{lemma}[\leanref{MobiusSeamChartableR2.lean}]
$\mathrm{unfoldFD}$ is $\sim$-invariant: if $p \sim q$ then $\mathrm{unfoldFD}(p) = \mathrm{unfoldFD}(q)$.
\end{lemma}

\begin{theorem}[\leanref{MobiusSeamChartableR2.lean}]
\label{thm:equator}
\AfterTheoremHead
\TheoremLeanTag{chartableR2\_mobiusQuotientMk\_of\_left\_seam\_half}
$\mathrm{ChartableR2}(\llbracket (0, \tfrac{1}{2}) \rrbracket)$.
\end{theorem}

\subsection{Assembly: All Interior Points Are ChartableR2}

\begin{theorem}[\leanref{MobiusSeamChartableR2.lean}]
\label{thm:interior-height}
\AfterTheoremHead
\TheoremLeanTag{chartableR2\_mobiusQuotientMk\_of\_interior\_height}
For every $p \in [0,1]^2$ with $0 < p_2 < 1$:
\[
  \mathrm{ChartableR2}(\llbracket p \rrbracket).
\]
\end{theorem}

\subsection{M\"obius C4: Boundary Biconditional}

\begin{theorem}[\leanref{MobiusChartableBoundary.lean}]
\label{thm:mobius-c4}
\AfterTheoremHead
\TheoremLeanTag{mobiusStripBoundary\_iff\_not\_chartableR2\_of\_forall\_off\_edge\_chartable}
For all $x \in \mathrm{MobiusStrip}$:
\[
  x \in \partial_{\mathrm{Mob}} \;\iff\; \neg\,\mathrm{ChartableR2}(x).
\]
\end{theorem}

% ============================================================================
% 6. M-FINAL: THE NON-HOMEOMORPHISM
% ============================================================================
\section{M-FINAL: The M\"obius Strip Is Not Homeomorphic to the Closed Cylinder}
\label{sec:mfinal}

\begin{theorem}[\leanref{MobiusSeamChartableR2.lean}, \leanref{ChartableR2Bridge.lean}]
\label{thm:mfinal}
\AfterTheoremHead
\TheoremLeanTag{mobiusStrip\_not\_homeomorphic\_closedCylinder}
\[
  \mathrm{IsEmpty}\!\left(\mathrm{MobiusStrip} \simeq_t \mathrm{ClosedCylinder}\right).
\]
\end{theorem}

\begin{proof}[Proof sketch]
Instantiate Theorem~\ref{thm:generic} with $b_X = \partial_{\mathrm{Mob}}$,
$b_Y = \partial_{\mathrm{Cyl}}$, the biconditionals
(Theorems~\ref{thm:cylinder-c4} and~\ref{thm:mobius-c4}), and the boundary
incompatibility (Theorem~\ref{thm:boundary-contrast}).
\end{proof}

% ============================================================================
% 7. THE EXTENDED MASTER THEOREM AND INSTANTIATIONS
% ============================================================================
\section{The Extended Master Theorem and Instantiations}
\label{sec:extended}

\subsection{Formal Statement}

\begin{theorem}[\leanref{Main.lean}]
\label{thm:extended}
\AfterTheoremHead
\TheoremLeanTag{representational\_regress\_master\_extended}
For any representational system $R$:
\[
  \mathrm{RepresentationalRegressMasterExtended}(R)
\]
where this abbreviation expands to:
\begin{enumerate}[label=(\roman*)]
\item $\mathrm{representational\_regress\_master\_claim}(R)$\newline
\noindent (regress + wall + $T_2$ transport),
\item $\mathrm{IsEmpty}(H^1 \simeq_t \mathbb{R}^1)$ (1D half-space vs.\ line;\newline
\noindent\leanref{representational\_regress\_topology\_halfLineModel}),
\item $\mathrm{IsEmpty}(\mathrm{MobiusStrip} \simeq_t \mathrm{ClosedCylinder})$ (M-FINAL).
\end{enumerate}
\end{theorem}

\noindent
This \emph{extends} \leanref{representational\_regress\_master}; it does not replace it.

\subsection{Bundled boundary models}

\begin{definition}[\leanref{ChartableR2Neighbor.lean}]
A \emph{\texttt{ChartableR2} boundary model} on $X$ is a subset $b_X \subseteq X$
together with a proof that
$\forall x,\; x \in b_X \iff \neg\,\mathrm{ChartableR2}(x)$.
\end{definition}

\noindent
\begin{itemize}[leftmargin=1.75em,itemsep=0.35ex,topsep=0.6ex,parsep=0pt]
\item The type \leanref{ChartableR2BoundaryModel} packages such data; the lemma
  \leanref{ChartableR2BoundaryModel.not\_homeomorphic} restates
  Theorem~\ref{thm:generic} in bundled form.
\item The development \leanref{ChartableR2ConcreteBoundaryModels} supplies\newline
  \leanref{mobiusStripChartableR2BoundaryModel} and
  \leanref{closedCylinderChartableR2BoundaryModel}.
\item The same-file theorem\newline
  \leanref{mobiusStrip\_closedCylinder\_boundary\_models\_not\_homeomorphic}\newline
  re-proves M-FINAL along that pipeline.
\end{itemize}

\subsection{Instantiations}
\label{sec:instantiations}

Theorems earlier in the paper are unconditional. This subsection records how the
same statements bear on four debates outside pure mathematics. Where an extra
premise is required, it is stated explicitly; nothing in the formalization
defaults a controversial premise to true.

\subsubsection{Consciousness and self-presence}

\paragraph{Philosophical bridge (premise P2).}
Some accounts distinguish representational self-reference from a stronger notion
of \emph{self-presence} (first-person givenness, non-representational
acquaintance, etc.). Premise~\textbf{P2} is \emph{not} a theorem in the library:
it says only that, for the purposes of a topological comparison, one class of
phenomena is modeled by a space globally akin to $\mathrm{MobiusStrip}$ (here:
one connected boundary band in the witness pair, quotient non-orientability in
the usual informal picture) and another class by a space akin to
$\mathrm{ClosedCylinder}$ (two disjoint boundary components, ``two-sided''
organization in the model).

\paragraph{What is proved.}
Theorem~\ref{thm:mfinal} is unconditional:
$\mathrm{MobiusStrip} \not\simeq_t \mathrm{ClosedCylinder}$. The proof
localizes ``boundary'' via \texttt{ChartableR2} and compares boundary
subspaces. Any other pair of compact spaces that admit the same pattern of
biconditionals and subtype incompatibility is also not homeomorphic, by
Theorem~\ref{thm:generic}.

\paragraph{Conditional consequence.}
If~(i)~one treats iterative representation as living on a substrate whose
\texttt{ChartableR2}-detected boundary matches the \emph{cylinder} side of the
development, and~(ii)~one accepts~P2 for a target phenomenon modeled on the
\emph{M\"obius} side, then no global homeomorphism can identify that substrate
with the target: the obstruction is topological. Under those assumptions,
resource limits (compute, context length, training data) address a different
question; they do not alter homeomorphism type. Whether~P2 is acceptable is a
separate philosophical issue; the formal work neither proves nor assumes it.

\subsubsection{Symbol grounding and semantic regress}

In the symbol-grounding problem, arbitrary symbols must be interpreted, and
interpreting the interpretation threatens a regress unless a non-representational
stop is posited. Formally, a representational system already contains the morphism
\texttt{represent} whose iterates index ``one more representational step.'' The
type alias \leanref{SymbolSystem} is definitionally identical to
\leanref{RepresentationalSystem}; lemmas such as
\leanref{symbolGrounding\_metaRegressArrow\_injective} record that distinct
natural-number depths correspond to \emph{distinct} endomorphisms of the
awareness object. Thus semantic ``levels of interpretation'' inherit the same
non-collapse facts as the categorical tower---not because the formalism defines
``meaning,'' but because the regress of \emph{structured} interpretation steps is
what \texttt{metaRegressArrow} tracks.

\subsubsection{Diagonal arguments and self-interpretation}

Lawvere's theorem in \texttt{Type} implies that there is no family
$s : A \to A \to B$ that realizes every unary $g : A \to B$ as $s(a)$ for some
parameter $a$, once $B$ carries an endomorphism without fixed points
(e.g.\ \leanref{no\_universal\_parametric\_unary\_nat} for $B = \mathbb{N}$,
or \leanref{no\_universal\_parametric\_unary\_bool} for $B = \mathrm{Bool}$
with $\lnot$---the decidable/computability analogue).

Crucially, the diagonal obstruction does not require the adversary's
self-model to be surjective.
\leanref{lawvere\_diagonal\_not\_in\_range} shows
that for \emph{any} $s : A \to A \to B$ and any fixed-point-free $f : B \to B$,
the diagonal function $a \mapsto f(s\,a\,a)$ is \emph{never} equal to $s\,a_0$
for any $a_0$. In other words, even a \emph{partial} self-interpreter
necessarily has a blind spot containing its own diagonal---the adversary
cannot circumvent the obstruction by abandoning totality.

\leanref{LawvereRegressBridge} collects such statements and connects them to the
categorical presentation: in particular,
\leanref{lawvere\_fixed\_point\_stays\_representational} says that for every
$n$ and every object witness $c$, the iterate
$\mathrm{OntologicalSlot.endo}(\mathrm{represent}^n)$ is unequal to
$\mathrm{OntologicalSlot.obj}(c)$. So however high the tower of self-applicative
coding rises inside the type discipline, the formalism still distinguishes the
morphism-slot tag from the global-element tag---a different statement from
Lawvere's pointwise fixed-point existence for $B \to B$, but the intended hinge
between diagonal phenomena and the \texttt{OntologicalSlot} reading of the wall.

\subsubsection{Quantum measurement and observer chains}

Talk of apparatus, environment, and Wigner's-friend hierarchies suggests a chain of
systems, each described relative to a larger ``observer.'' If one ever gives such
a chain a \emph{topological} state space $\mathrm{Obs}$, together with a subset
$\partial_{\mathrm{obs}}$ such that membership agrees with
$\neg\,\mathrm{ChartableR2}$ as in the cylinder/M\"obius story, and if one shows
that $\partial_{\mathrm{obs}}$ is not homeomorphic to
$\mathrm{mobiusStripBoundary}$ as a subspace, then
Theorem~\ref{thm:generic} applies immediately:
\leanref{QuantumObserverChainHypothesis} is only this data packaged as
assumptions, and \leanref{QuantumObserverChainHypothesis.not\_homeomorphic\_mobiusStrip}
is the conclusion $\mathrm{IsEmpty}(\mathrm{Obs} \simeq_t \mathrm{MobiusStrip})$.
The present library does \emph{not} define Hilbert spaces, observables, or
projection postulates; any use of the hypothesis structure in physics requires an
independent argument that the modeled readout topology satisfies its fields.

\subsection{Scaling of language models and similar systems}

Training larger models extends a representational architecture; it does not, by
itself, produce a homeomorphism between a given topological model of that
architecture and a different global model. If the relevant state space is
modeled with cylinder-type boundary data and a target phenomenon demands
M\"obius-type data, Theorem~\ref{thm:generic} encodes the sense in which the gap
is not closed by parameter count or depth alone.

\subsection{Relationship to the Trans-Computational Processing argument}

An earlier line of work argues from computability---halting,
G\"odel-style diagonalization, and related limits---to constraints on
``self-contained'' computational mind models. The topological strand here is
independent: it turns on \texttt{ChartableR2} sensors and boundary
subspaces, not on recursiveness. The two approaches can reinforce a common
\emph{limitative} conclusion about representation closed under its own rules,
but neither formally entails the other.

\subsection{No-escape analysis: horn-by-horn closure}
\label{sec:noescape}

An adversary confronting the representational regress has, at first glance,
several avenues of retreat.  Each is formalized and shown to fail.  The
numbering below follows the order in which the escape was identified; each
``horn'' names an adversarial move, and the closing theorem blocks it.

\paragraph{Horn~1: Infinite injective tower.}
\emph{Move:} accept $\mathrm{iter\_injective}$ and try to use the tower itself
as a self-representation resource.
\emph{Closure:} Theorem~\ref{thm:regress} yields infinitely many distinct
meta-levels;
Theorem~\ref{thm:wall} (the Lawvere wall) ensures the endo/obj distinction
persists at every level.  The tower does not collapse into closure.

\paragraph{Horn~2: Finite-order / periodic iterates.}
\emph{Move:} weaken $\mathrm{iter\_injective}$---let
$\mathrm{represent}^n = \mathrm{represent}^m$ for some $n \neq m$, so the
loop has only finitely many distinct powers.\newline
\noindent\emph{Closure:}
\leanref{IterateClosure.End.pow\_iterate\_dichotomy}
proves the dichotomy: either the iterates are all distinct (Horn~1) or
there exists a bound $B$ such that every power equals some $e^i$ with
$i < B$.  In the finite case the system's representational depth is
bounded; it cannot host an injective tower at all.

\paragraph{Horn~3: Cross-object readout via section--retraction.}
\emph{Move:} represent $A$ \emph{inside} a different object $B$ using a
section $s : A \to B$ and retraction $r : B \to A$ with $s \circ r = \mathrm{id}_A$.
\emph{Closure:}
\leanref{CrossObjectRepresentation.section\_retraction\_no\_injective\_tower\_either\_side}
shows that $r \circ s$ is idempotent on $B$ (so $(r \circ s)^1 = (r \circ s)^2$,
blocking injective powers on the host side) and $s \circ r = \mathrm{id}_A$
(so all powers are the identity on the source side---trivially periodic).
Neither side supports an injective $\mathbb{N}$-tower.

\paragraph{Horn~4: Partial (non-surjective) self-model.}
\emph{Move:} concede that no \emph{total} self-interpreter exists, but claim
a useful \emph{partial} one.
\emph{Closure:}
\leanref{lawvere\_diagonal\_not\_in\_range}
requires no surjectivity of $s$: for \emph{any} $s : A \to A \to B$ and any
fixed-point-free $f : B \to B$, the diagonal function $a \mapsto f(s\,a\,a)$
is never in the range of $s$.  Every partial self-model necessarily has a
blind spot that includes its own diagonal---the information that matters most
for self-capture is exactly the information it cannot represent.

\paragraph{Horn~5: Different topological models.}
\emph{Move:} reject the cylinder/M\"obius witness; use a different geometric
model.
\emph{Closure:} Theorem~\ref{thm:generic}
(\leanref{not\_homeomorphic\_of\_chartableR2\_boundary\_contrast})
is \emph{model-agnostic}: it applies to \emph{any} pair of spaces whose
$\mathrm{ChartableR2}$-co-characterized boundaries are subtype-incompatible.
The cylinder and M\"obius strip are a discharged witness;
any other pair meeting the same three hypotheses imports the same conclusion.

\paragraph{Horn~6: Computability / decidable self-reference.}
\emph{Move:} restrict to decidable predicates or finite-valued functions,
where ``self-reference'' seems more tractable than Cantorian diagonals.
\emph{Closure:}
\leanref{no\_universal\_parametric\_unary\_bool}
shows that no $s : A \to A \to \mathrm{Bool}$ can realize every
$g : A \to \mathrm{Bool}$, since $\lnot : \mathrm{Bool} \to \mathrm{Bool}$
is fixed-point-free.  The obstruction is not an artifact of infinite
cardinality; it holds in the decidable, two-valued setting.

\paragraph{Residual openings (not closed here).}
Two further routes are not formally addressed in this paper:
\emph{(i)}~\emph{higher-categorical self-maps} (a self-representation
living in a $2$-category or $\infty$-category, where the monoid of
endomorphisms is replaced by a richer hom-object); and
\emph{(ii)}~\emph{non-categorical frameworks} (a system that resists
categorical encoding entirely).  The first is subject to analogous
monoid-element arguments once a truncation to the $1$-categorical
endomorphism monoid is performed; the second falls outside the scope of
any categorical formalization by definition.  Both are noted as
limitations in \S\ref{sec:limits}.

\subsection{Admissibility: what the closures tell neighboring debates}
\label{sec:admissibility}

Given that Horns~1--6 are closed, the formal landscape constrains what
\emph{admissible} solutions look like in each neighboring debate.  The
reasoning in every case is the same: the closures establish a structural
barrier, so any proposed solution must either (a)~operate outside the
barrier's hypotheses, or (b)~accept the consequences.

\paragraph{Consciousness and self-presence.}
If a system's state-space has cylinder-type boundary organization (two
disjoint boundary components, each $\neg\mathrm{ChartableR2}$), then no
continuous global identification (homeomorphism) maps it to a M\"obius-type
target (one connected boundary band).  An admissible account of self-presence
must therefore either
(i)~posit that the relevant substrate \emph{already} has the target topological
type (in which case it was never purely representational in the cylinder
sense), or
(ii)~invoke a discontinuous or non-topological bridge (which the formalism
does not model and which would require independent justification).
Scaling---more parameters, longer context windows, deeper architectures---extends
a representational system \emph{within} a fixed topological type; it does not
change boundary organization.

\paragraph{Symbol grounding.}
Any symbol system that iterates interpretation is caught in the
Horns~1/2 dichotomy: either infinite distinct levels (full regress plus
Lawvere wall) or finite-bounded depth (finite loop, no injective tower).
Horn~4 adds that even a \emph{partial} interpretation function misses its
own diagonal.  Admissible grounding must therefore come from
\emph{outside} the representational loop: the system cannot ground itself
by iterating its own interpretive machinery, whether totally or partially.

\paragraph{Diagonal arguments / AI self-interpretation.}
Any proposed self-interpreter $s : A \to A \to B$ (total or partial)
necessarily fails to include its own diagonal
(\leanref{lawvere\_diagonal\_not\_in\_range}); this holds for
$B = \mathbb{N}$ (numeric outputs), $B = \mathrm{Bool}$ (decidable
predicates), and any $B$ with a fixed-point-free endomorphism.
Admissible self-interpreters \emph{must} be incomplete in a precise
sense: they cannot model their own diagonal behavior.  This is not a
resource limitation but a structural one; no amount of compute, training
data, or architectural ingenuity changes the diagonal exclusion.

\paragraph{Quantum measurement chains.}
If the measurement chain's state space has a boundary pattern matching the
cylinder side of the development, then no homeomorphism identifies it with a
M\"obius-type target.  The chain must either introduce genuinely new
topological structure at some level---not representational iteration of the
same type---or accept that its boundary organization differs from the target
pattern.

\paragraph{LLM and AI scaling.}
Scaling extends a representational architecture within a fixed topological
type. If that type has cylinder-side boundary organization, scaling does not
produce a homeomorphism to a M\"obius-type target.
Admissible paths to a different topological type require
\emph{structural} change (different topology of the underlying state
space), not quantitative extension (more of the same).

% ============================================================================
% 8. LIMITATIONS AND SCOPE
% ============================================================================
\section{Limitations and scope}
\label{sec:limits}

\begin{enumerate}
\item \textbf{Hypothesis, not derivation.} $\mathrm{iter\_injective}$ is bundled,
  not proved from minimal data. Not every $f : A \to A$ in every category satisfies it.
  The regress is conditional on this non-triviality hypothesis.

\item \textbf{Lawvere in \texttt{Type}, not abstract CCC.} The formalization uses
  $\mathrm{Type}\ u$ and\newline
  $\mathrm{MonoidalClosed}(\mathrm{Type}\ u)$, not a general closed category over arbitrary~$C$.
  A fully abstract CCC version remains future work.

\item \textbf{\texttt{OntologicalSlot} is definitional, not metaphysical.} The disjointness $\mathrm{endo} \neq \mathrm{obj}$ is trivially true by constructor discrimination. This is the point, but the triviality means the result is \emph{semantic} (about the structure of type theory) rather than a deep theorem.

\item \textbf{Models are chosen, not unique.} The closed cylinder and M\"obius strip are \emph{specific} topological models chosen for clarity and tractability. A different philosophical account might use different geometric witnesses (e.g., Klein bottle, different quotient). The argument works for any two models separated by a topological invariant; these two are the cleanest.

\item \textbf{P2 is philosophical.} The premise ``consciousness requires non-orientable topology'' is not a theorem. Section~\ref{sec:instantiations} states only what follows when P2 is adjoined to the proved non-homeomorphism; accepting or rejecting P2 remains an independent issue.

\item \textbf{No $\mathrm{IsManifold}$ polish.} The M\"obius strip is not given a full Mathlib $\mathrm{IsManifold}$ instance. The proof uses $\mathrm{ChartableR2}$ (open neighborhoods homeomorphic to $\mathbb{R}^2$) as a self-contained predicate, sidestepping the need for a full manifold-with-boundary atlas. This is sufficient for the non-homeomorphism but less elegant than a full atlas approach.

\item \textbf{Route W is incomplete.} An alternative proof via circle-winding / doubling obstructions (Steps~1--4 in \leanref{CylinderMobiusNonhomeo.lean}) is partially formalized. Step~5 (transport under a hypothetical global homeomorphism) is not proved. The flagship proof via $\mathrm{ChartableR2}$ is complete and does not depend on Route~W.

\item \textbf{\texttt{QuantumObserverChainHypothesis}.} This definition only
  bundles topological assumptions (boundary set, \texttt{ChartableR2}
  biconditional,\newline subtype non-homeomorphism to the M\"obius boundary). It does not
  import Hilbert spaces, operators, or measurement axioms; a physical
  application would need a separate argument that a concrete model satisfies
  those fields.

\item \textbf{Partial self-models: blind spot, not total impossibility.}
  \leanref{lawvere\_diagonal\_not\_in\_range} proves that every partial
  self-model misses its own diagonal; it does not prove that partial
  self-models are \emph{useless}---only that they are necessarily incomplete
  in a precise, diagonal-shaped way.  Whether that incompleteness matters
  for a given application is a separate question.

\item \textbf{Higher-categorical and non-categorical frameworks.}
  The horn closures rely on $1$-categorical endomorphism monoids.  A
  self-representation living in a $2$-category or $\infty$-category
  replaces the monoid with a richer hom-object; the present arguments
  apply after truncation to the underlying $1$-categorical monoid, but
  a native higher-categorical treatment is not given.  Systems that
  resist categorical encoding entirely fall outside any categorical
  formalization by definition.

\item \textbf{Computability: decidable analogue only.}
  \leanref{no\_universal\_parametric\_unary\_bool} closes the
  $\mathrm{Bool}$-valued Lawvere horn.  A full treatment of Turing-degree
  obstructions (Rice's theorem, Kleene's recursion theorem) in the same
  framework is not attempted; the present result is the finite/decidable
  endpoint of the diagonal family.
\end{enumerate}

% ============================================================================
% 9. RELATED WORK
% ============================================================================
\section{Related work}
\label{sec:related}

\subsection{Formalized topology}
Mathlib's manifold and topology libraries underpin the half-space and chart
lemmas; other proof assistants also host substantial topology. The weight here is
on a concrete non-homeomorphism coupled to representational-system infrastructure,
not on novel classification of differential invariants.

\subsection{Lawvere's theorem in proof theory}
Lawvere's fixed-point argument appears in multiple formal developments.
\texttt{OntologicalSlot} and the wall lemmas supply complementary
category-internal tagging alongside the \texttt{Type}-level diagonal theorem.

\subsection{Consciousness, G\"odel, and AI}
G\"odel- and computability-based limits on mechanism (including Penrose-style
arguments, discussions of strange loops and self-reference in the sense of
Hofstadter, and information-theoretic theories of
consciousness) occupy a large literature orthogonal to the cylinder/M\"obius
homeomorphism question, though informal exposition sometimes elides the
difference.

\subsection{Topology in philosophy of mind}
Topological figures of speech appear in some philosophy of mind. The present
work differs in that its central topological claims are proved and the
generalization pattern is explicit.

% ============================================================================
% 10. CONCLUSION
% ============================================================================
\section{Conclusion}
\label{sec:conclusion}

The formalization delivers three interlocking layers:

\begin{enumerate}[label=(\roman*)]
\item A machine-checked regress and Lawvere infrastructure (categorical strand),
\item A general $\mathrm{ChartableR2}$ boundary-contrast lemma with a fully
  discharged compact witness
  $\mathrm{MobiusStrip} \not\simeq_t \mathrm{ClosedCylinder}$
  (topological strand), and
\item A systematic no-escape analysis (Horns~1--6,
  \S\ref{sec:noescape}) that formally closes
  the principal adversarial retreat routes: infinite regress,
  finite-order collapse, cross-object readout, partial self-models,
  model substitution, and decidable self-reference.
\end{enumerate}

\noindent
The central insight is not that infinite regress is bad---that is widely
acknowledged.  The insight is that the barrier is \textbf{topological and
structural}, not computational or resource-bound.  The obstacle is
\emph{wrong type}, not \emph{inexhaustible queue}.
A homeomorphism obstruction is a finite-data invariant: two spaces are
either homeomorphic or not, regardless of how many modeling steps one
enumerates.  No amount of compute, no deeper self-model, no richer
architecture changes the boundary pattern of a topological type.  And the
Lawvere diagonal exclusion
(\leanref{lawvere\_diagonal\_not\_in\_range})
shows that even \emph{partial} self-models necessarily have a blind spot
containing their own diagonal---the very information needed for
self-capture is the information they cannot represent.

\paragraph{What is not being claimed.}
The theorems are unconditional mathematics; the \emph{application} to
any particular domain requires domain-specific premises (P2 for
consciousness, \texttt{ChartableR2} biconditionals for a specific
state-space model, etc.).  The formalism does not prove that
consciousness is non-representational, that AI cannot be conscious, or
that symbol grounding is impossible.  It proves that \emph{if} the
relevant state space has the boundary pattern characterized here,
\emph{then} a homeomorphism to a space with a different boundary
pattern does not exist---and that the standard adversarial attempts to
work around that barrier fail in formally checkable ways.

\paragraph{The admissibility picture.}
The horn closures (\S\ref{sec:noescape}) together with the
admissibility analysis (\S\ref{sec:admissibility}) leave a clean
landscape.  In each neighboring debate---consciousness,
symbol grounding, diagonal limitations, measurement chains, AI
scaling---the admissible solutions share a common shape:
\emph{structural change} to the topological type of the substrate,
not \emph{quantitative extension} within a fixed type.  The formalism
makes this precise and machine-checkable, so that future claims about
circumventing the barrier can be evaluated against a fixed standard.

\medskip
\noindent\textbf{Future work.} A full $\mathrm{IsManifold}$ instance for
$\mathrm{MobiusStrip}$; completion of the optional Route~W winding track; a more
abstract Lawvere statement for general cartesian closed categories; native
higher-categorical horn closures (2-category and $\infty$-category analogues);
further models where $\mathrm{ChartableR2}$ boundary data can be established;
and a full Turing-degree analogue of the decidable Lawvere horn.

% ============================================================================
% APPENDICES
% ============================================================================
\appendix

\section{Lean Module Map}
\label{app:modules}

\begin{table}[h]
\centering
\footnotesize
\setlength{\tabcolsep}{3pt}
\begin{tabular}{@{}>{\raggedright\arraybackslash}p{3.95cm}>{\raggedright\arraybackslash}p{11.45cm}@{}}
\toprule
\textbf{Module} & \textbf{Role} \\
\midrule
\leanref{Basic} & \leanref{RepresentationalSystem}, iterates, \leanref{OntologicalSlot} \\
\leanref{Regress} & Non-termination, unbounded levels \\
\leanref{FixedPoints} & \leanref{EndoVsPoint}, Lawvere wall non-dissolution \\
\leanref{Lawvere} & Fixed-point theorem in \texttt{Type} \\
\leanref{LawvereCCCType} & \texttt{MonoidalClosed} packaging \\
\leanref{Orientability} & $T_2$ homeomorphism invariance \\
\leanref{CylinderMobius} & Fundamental domain, quotient, compact Hausdorff, boundary band \\
\leanref{CylinderBoundary} & Closed cylinder boundary: two disjoint connected faces \\
\leanref{CylinderChartableBoundary} & Cylinder C4 biconditional \\
\leanref{ChartableR2Neighbor} & \leanref{ChartableR2}, \leanref{ChartableR2BoundaryModel}; openness / stability \\
\leanref{ChartableR2Bridge} & \leanref{Homeomorph.chartableR2\_iff}; \leanref{not\_homeomorphic\_of\_chartableR2\_boundary\_contrast}; \leanref{ChartableR2BoundaryModel.not\_homeomorphic} \\
\leanref{ChartableR2ConcreteBoundaryModels} & M\"obius/cylinder boundary models; conditional observer chain; \leanref{QuantumObserverChainHypothesis} \\
\leanref{SymbolGrounding} & \leanref{SymbolSystem}; \leanref{symbolGrounding\_*} \\
\leanref{LawvereRegressBridge} & \leanref{no\_universal\_parametric\_unary\_nat}, \leanref{no\_universal\_parametric\_unary\_bool}; \leanref{lawvere\_diagonal\_not\_in\_range}; \leanref{lawvere\_fixed\_point\_stays\_representational} \\
\leanref{NoEscapeClosure} & Iterate collisions vs.\ injective towers; section--retraction idempotent readout \\
\leanref{ChartableR2Models} & Interior charts; cylinder interior \leanref{ChartableR2} \\
\leanref{MobiusChartableBoundary} & Open-cell charts; boundary $\neg$\leanref{ChartableR2}; M\"obius C4 packaging \\
\leanref{MobiusSeamChartable} & Saturated seam patches (open, saturated) \\
\leanref{MobiusSeamChart} & IFT-based local model; trig coordinates \\
\leanref{MobiusSeamTrigInject} & Trig injectivity on quotient windows \\
\leanref{MobiusSeamChartableR2} & Seam charts (trig + unfold); interior-height; \textbf{M-FINAL} \\
\leanref{MobiusCylinderBoundaryContrast} & Boundary subspaces not homeomorphic \\
\leanref{CylinderMobiusNonhomeo} & Route W (optional; Steps 1--4) \\
\leanref{HalfLineVsLine} & 1D half-space $\neq$ line \\
\leanref{HalfPlaneVsPlane} & 2D half-space $\neq$ plane \\
\leanref{HalfSpaceNeighborVsPlane} & Local half-space $\neq$ open plane patch \\
\leanref{PuncturedPlaneNotSimplyConnected} & $\mathbb{C} \setminus \{0\}$ not simply connected \\
\leanref{OpenCompactChartObstruction} & Compact $\mathbb{R}^2$ patch $\neq$ $\mathbb{R}^2$ \\
\leanref{Main} & Master theorem + extended bundle \\
\bottomrule
\end{tabular}
\end{table}

\section{Key Theorem Index}
\label{app:theorems}

\begin{table}[h]
\centering
\footnotesize
\setlength{\tabcolsep}{3pt}
\begin{tabular}{@{}>{\raggedright\arraybackslash}p{3.85cm}>{\raggedright\arraybackslash}p{7.55cm}>{\raggedright\arraybackslash}p{3.15cm}@{}}
\toprule
\textbf{Lean name} & \textbf{Statement} & \textbf{Module} \\
\midrule
\leanref{regress\_iterates\_unbounded} & $\forall B, \exists L > B$, distinct iterate & \leanref{Regress} \\
\leanref{lawvere\_fixed\_point\_Type} & Universal enumerator $\Rightarrow$ every endo has fixed point & \leanref{Lawvere} \\
\leanref{lawvere\_wall\_is\_not\_dissolution} & $\mathrm{endo}(\mathrm{represent}) \neq \mathrm{obj}(\mathrm{fp})$ & \leanref{FixedPoints} \\
\leanref{instT2SpaceMobiusStrip} & M\"obius strip is $T_2$ & \leanref{CylinderMobius} \\
\leanref{isConnected\_mobiusStripBoundary} & Boundary band is connected & \leanref{CylinderMobius} \\
\leanref{closedCylinder\_boundary\_eq\_union} & Cylinder boundary = 2 disjoint faces & \leanref{CylinderBoundary} \\
\leanref{euclideanHalfSpace2\_not\_homeomorphic\_euclidean2} & $H^2 \not\cong_t \mathbb{R}^2$ & \leanref{HalfPlaneVsPlane} \\
\leanref{not\_homeomorphic\_of\_chartableR2\_boundary\_contrast} & Generic: incompatible boundaries $\Rightarrow$ $\neg$ homeo & \leanref{ChartableR2Bridge} \\
\leanref{closedCylinder\_boundaryUnion\_iff\_not\_chartableR2} & Cylinder C4 biconditional & \leanref{CylinderChartableBoundary} \\
\leanref{not\_chartableR2\_of\_mem\_mobiusStripBoundary} & M\"obius boundary $\Rightarrow$ $\neg$\leanref{ChartableR2} & \leanref{MobiusChartableBdy} \\
\leanref{chartableR2\_mobiusQuotientMk\_of\_interior\_height} & Interior height $\Rightarrow$ \leanref{ChartableR2} & \leanref{MobiusSeamChartableR2} \\
\leanref{mobiusStripBoundary\_not\_homeomorphic\ldots} & Boundary subtypes not homeo & \leanref{MobiusCylBdyContrast} \\
\leanref{mobiusStrip\_not\_homeomorphic\_closedCylinder} & \textbf{M-FINAL} & \leanref{MobiusSeamChartableR2} \\
\leanref{representational\_regress\_master} & Core master conjunction & \leanref{Main} \\
\leanref{representational\_regress\_master\_extended} & Extended bundle (+ topology) & \leanref{Main} \\
\leanref{QuantumObserverChainHypothesis.not\_homeomorphic\_mobiusStrip} & Conditional: obs.\ chain $\not\simeq_t$ M\"obius & \leanref{ChartableR2ConcreteBoundaryModels} \\
\leanref{no\_universal\_parametric\_unary\_nat} & No universal $s : A \!\to\! A \!\to\! \mathbb{N}$ for all $g : A \!\to\! \mathbb{N}$ & \leanref{LawvereRegressBridge} \\
\leanref{IterateClosure.End.pow\_iterate\_dichotomy} & Injective iterate powers vs.\ finite index bound & \leanref{NoEscapeClosure} \\
\leanref{CrossObjectRepresentation.section\_retraction\_no\_injective\_tower\_either\_side} & Both-side: $\neg$ injective $\mathbb{N}$-powers on source \emph{or} host & \leanref{NoEscapeClosure} \\
\leanref{lawvere\_diagonal\_not\_in\_range} & Diagonal never in range of any $s : A \!\to\! A \!\to\! B$ (no surjectivity) & \leanref{LawvereRegressBridge} \\
\leanref{no\_universal\_parametric\_unary\_bool} & No universal $s : A \!\to\! A \!\to\! \mathrm{Bool}$ for all $g : A \!\to\! \mathrm{Bool}$ & \leanref{LawvereRegressBridge} \\
\bottomrule
\end{tabular}
\end{table}

\section{Reproduction}
\label{app:repro}

\begin{verbatim}
cd representational-regress-lean
lake update
lake exe cache get
lake build RepresentationalRegress
# Verify: 0 sorry
rg "sorry" RepresentationalRegress/
\end{verbatim}

\noindent
Toolchain: \texttt{leanprover/lean4:v4.29.0-rc6}. Mathlib: \texttt{v4.29.0-rc6}.

\end{document}
